---
title: "Hi-C Cscore Compartment Analysis – REH samples"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: hide
    fig_width: 8
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)
knitr::opts_knit$set(root.dir = "R:/R_STIK/Hi-C/HiC_marta_Oct25/cscore")
```

---

# Overview

This notebook analyses **A/B chromatin compartments** inferred from Hi-C data using
Cscore bedgraph files for three conditions:

| Condition | Description | Files |
|-----------|-------------|-------|
| **REH** | Parental cell line | `REH_tp.bedgraph_correct.bedgraph` |
| **EP1** | EP1 treatment | `HiC-REH-EP1-{1,2}_tp…` + merged |
| **EP1-Aro** | EP1 + Aromatic treatment | `HiC-REH-EP1-Aro-{1,2}_tp…` + merged |

**Analysis steps:**

1. Load & merge all bedgraph files  
2. QC: PCA and Spearman correlation across replicates and merged samples  
3. A/B compartment classification  
4. Compartment switching dynamics  
5. Heatmap of cscore values  

---

# 1. Libraries and output directories

```{r libraries}
library(dplyr)
library(ggplot2)
library(reshape2)
library(ggrepel)
library(pheatmap)
library(viridis)
library(tidyr)
library(grid)
library(cowplot)

output_dir <- "Comp_Analysis"

files_dir  <- file.path(output_dir, "files")
plots_dir  <- file.path(output_dir, "plots")

for (d in c(files_dir, plots_dir)) {
  if (!dir.exists(d)) dir.create(d, recursive = TRUE)
}
```

---

# 2. Data loading

All bedgraph files are loaded from the cscore iteration trial directory.
Each file contains four columns: `chr`, `start`, `end`, and the cscore `score`.

```{r load-data}
data_dir <- "R:/R_STIK/Hi-C/HiC_marta_Oct25/cscore"

read_cscore <- function(filename, score_colname) {
  read.table(file.path(data_dir, filename),
             sep       = "\t",
             col.names = c("chr", "start", "end", score_colname))
}

# Individual replicates
REH        <- read_cscore("REH_tp.bedgraph_correct.bedgraph",           "score_REH")
EP1_1      <- read_cscore("HiC-REH-EP1-1_tp.bedgraph_correct.bedgraph", "score_EP1_1")
EP1_2      <- read_cscore("HiC-REH-EP1-2_tp.bedgraph_correct.bedgraph", "score_EP1_2")
EP1_Aro_1  <- read_cscore("HiC-REH-EP1-Aro-1_tp.bedgraph_correct.bedgraph", "score_EP1_Aro_1")
EP1_Aro_2  <- read_cscore("HiC-REH-EP1-Aro-2_tp.bedgraph_correct.bedgraph", "score_EP1_Aro_2")

# Merged replicates
EP1_merge     <- read_cscore("HiC-REH-EP1_tp.bedgraph_correct.bedgraph",     "score_EP1_merge")
EP1_Aro_merge <- read_cscore("HiC-REH-EP1-Aro_tp.bedgraph_correct.bedgraph", "score_EP1_Aro_merge")
```

## 2.1 Merge into a single dataframe

All files are merged by genomic coordinates (`chr`, `start`, `end`).

```{r merge-data}
cs_REH <- Reduce(
  function(x, y) merge(x, y, by = c("chr", "start", "end")),
  list(REH, EP1_1, EP1_2, EP1_Aro_1, EP1_Aro_2, EP1_merge, EP1_Aro_merge)
)

cat("Columns:", paste(colnames(cs_REH), collapse = ", "), "\n")

head(cs_REH)
```

---

# 3. Quality Control

## 3.1 PCA

PCA is run on the transposed cscore matrix (samples as rows, genomic bins as
columns) to assess whether:

- Replicates of the same condition cluster tightly together  
- Merged samples fall close to their corresponding replicates  
- The three conditions separate along the principal components  

**Shape legend:** circles = parental REH, squares = individual replicates, triangles = merged.  
**Colour legend:** condition.

```{r pca}
sample_labels <- c("REH", "EP1_1", "EP1_2", "EP1_Aro_1", "EP1_Aro_2",
                   "EP1_merge", "EP1_Aro_merge")

cscore_matrix           <- as.matrix(cs_REH[, 4:10])
colnames(cscore_matrix) <- sample_labels
cscore_matrix           <- na.omit(cscore_matrix)

# remove bins with zero variance
cscore_matrix <- cscore_matrix[apply(cscore_matrix, 1, var) != 0, ]

# PCA
PCA_REH <- prcomp(t(cscore_matrix), scale. = TRUE)
pct_var <- round(100 * PCA_REH$sdev^2 / sum(PCA_REH$sdev^2), 1)

# Build plot dataframe
pca_df            <- as.data.frame(PCA_REH$x)
pca_df$sample     <- sample_labels
pca_df$condition  <- c("REH", "EP1", "EP1", "EP1_Aro", "EP1_Aro", "EP1", "EP1_Aro")
pca_df$type       <- c("original", "replicate", "replicate",
                        "replicate", "replicate", "merge", "merge")

# Colour and shape scales
cond_colors <- c("REH"     = "#e74c3c",
                 "EP1"     = "#3498db",
                 "EP1_Aro" = "#2ecc71")

type_shapes <- c("original"  = 21,
                 "replicate" = 22,
                 "merge"     = 24)

pca_before <- ggplot(pca_df, aes(x = PC1, y = PC2,
                   fill  = condition,
                   shape = type,
                   label = sample)) +
  geom_point(color = "black", stroke = 1.3, size = 5) +
  geom_text_repel(size = 4, max.overlaps = 20) +
  scale_fill_manual(values = cond_colors) +
  scale_shape_manual(values = type_shapes) +
  theme_bw() +
  labs(
    title = "Cscore PCA – REH samples (replicates + merges)",
    x     = paste0("PC1 (", pct_var[1], "% variance)"),
    y     = paste0("PC2 (", pct_var[2], "% variance)")
  ) +
  theme(
    panel.grid   = element_blank(),
    plot.title   = element_text(size = 15, face = "bold", hjust = 0.5),
    axis.title   = element_text(size = 13),
    axis.text    = element_text(size = 11),
    legend.title = element_text(size = 11),
    legend.text  = element_text(size = 10)
  ) +
  guides(
    fill  = guide_legend(title = "Condition", override.aes = list(shape = 21)),
    shape = guide_legend(title = "Type")
  )
print(pca_before)
ggsave(file.path(plots_dir, "pca_before_correction.pdf"), pca_before, width = 10, height = 6)
```
Now we remove some problematic regions that we detected through IGV visualization of the bedgraphs

```{r Problematic regions removal + PCA after correction}
# Regions to remove
chr4_start  <- 127470171
chr4_end    <- 141890371

chr21_start <- 4514218
chr21_end   <- 12001870

# Function to correct a bedgraph
correct_bedgraph <- function(df, score_col, flip_chr13 = FALSE){

  # remove chr4 problematic region
  df <- df[!(df$chr == "chr4" &
             df$end   >= chr4_start &
             df$start <= chr4_end), ]

  # remove chr21 problematic region
  df <- df[!(df$chr == "chr21" &
             df$end   >= chr21_start &
             df$start <= chr21_end), ]

  # flip chr13 if required
  if(flip_chr13){
    df[df$chr == "chr13", score_col] <- -df[df$chr == "chr13", score_col]
  }

  return(df)
}

# List all samples
samples <- list(
  REH            = list(df = REH,           col = "score_REH",            flip = TRUE),
  EP1_1          = list(df = EP1_1,         col = "score_EP1_1",          flip = TRUE),
  EP1_2          = list(df = EP1_2,         col = "score_EP1_2",          flip = FALSE),
  EP1_Aro_1      = list(df = EP1_Aro_1,     col = "score_EP1_Aro_1",      flip = TRUE),
  EP1_Aro_2      = list(df = EP1_Aro_2,     col = "score_EP1_Aro_2",      flip = TRUE),
  EP1_merge      = list(df = EP1_merge,     col = "score_EP1_merge",      flip = TRUE),
  EP1_Aro_merge  = list(df = EP1_Aro_merge, col = "score_EP1_Aro_merge",  flip = TRUE)
)

# Apply correction to all samples
corrected <- list()

for(name in names(samples)){

  sample_info <- samples[[name]]

  df_corr <- correct_bedgraph(
    sample_info$df,
    sample_info$col,
    sample_info$flip
  )

  # store corrected dataframe
  corrected[[name]] <- df_corr

  outfile <- paste0(name, "_manual_corrected.bedgraph")
  
  write.table(
    df_corr,
    outfile,
    sep = "\t",
    quote = FALSE,
    row.names = FALSE,
    col.names = FALSE
  )

  cat("Saved:", outfile, "\n")
}
```

### Re-merge corrected samples

```{r merge-corrected}
cs_REH_corr <- Reduce(
  function(x, y) merge(x, y, by = c("chr", "start", "end")),
  list(corrected$REH,
       corrected$EP1_1,
       corrected$EP1_2,
       corrected$EP1_Aro_1,
       corrected$EP1_Aro_2,
       corrected$EP1_merge,
       corrected$EP1_Aro_merge)
)

cat("Bins in corrected merged dataframe:", nrow(cs_REH_corr), "\n")
```

## 3.4 PCA (post-correction)

Re-run the same PCA on the corrected data. Samples should now cluster more
cleanly by condition, with replicates tighter and the chr4/chr13 artefacts
no longer driving separation.

```{r pca-corrected}
cscore_matrix_corr           <- as.matrix(cs_REH_corr[, 4:10])
colnames(cscore_matrix_corr) <- sample_labels
cscore_matrix_corr           <- na.omit(cscore_matrix_corr)
cscore_matrix_corr           <- cscore_matrix_corr[apply(cscore_matrix_corr, 1, var) != 0, ]

cat("Bins used for post-correction PCA:", nrow(cscore_matrix_corr), "\n")

PCA_corr <- prcomp(t(cscore_matrix_corr), scale. = TRUE)
pct_var_corr <- round(100 * PCA_corr$sdev^2 / sum(PCA_corr$sdev^2), 1)

pca_df_corr            <- as.data.frame(PCA_corr$x)
pca_df_corr$sample     <- sample_labels
pca_df_corr$condition  <- c("REH", "EP1", "EP1", "EP1_Aro", "EP1_Aro", "EP1", "EP1_Aro")
pca_df_corr$type       <- c("original", "replicate", "replicate",
                             "replicate", "replicate", "merge", "merge")

pca_after <- ggplot(pca_df_corr, aes(x = PC1, y = PC2,
                        fill  = condition,
                        shape = type,
                        label = sample)) +
  geom_point(color = "black", stroke = 1.3, size = 5) +
  geom_text_repel(size = 4, max.overlaps = 20) +
  scale_fill_manual(values = cond_colors) +
  scale_shape_manual(values = type_shapes) +
  theme_bw() +
  labs(
    title = "Cscore PCA – post-correction (replicates + merges)",
    x     = paste0("PC1 (", pct_var_corr[1], "% variance)"),
    y     = paste0("PC2 (", pct_var_corr[2], "% variance)")
  ) +
  theme(
    panel.grid   = element_blank(),
    plot.title   = element_text(size = 15, face = "bold", hjust = 0.5),
    axis.title   = element_text(size = 13),
    axis.text    = element_text(size = 11),
    legend.title = element_text(size = 11),
    legend.text  = element_text(size = 10)
  ) +
  guides(
    fill  = guide_legend(title = "Condition", override.aes = list(shape = 21)),
    shape = guide_legend(title = "Type")
  )
print(pca_after)
ggsave(file.path(plots_dir, "pca_after_correction.pdf"), pca_after, width = 10, height = 6)
```

## 3.2 Spearman correlation matrix

Pairwise Spearman correlation between all samples. Values close to 1 between
replicates of the same condition indicate good reproducibility.

```{r correlation}
cor_mat <- cor(cscore_matrix, method = "spearman")

spear_p <- ggplot(melt(cor_mat), aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  geom_text(aes(label = round(value, 2)), size = 3.2) +
  coord_equal() +
  theme_bw() +
  theme(
    axis.text.x  = element_text(angle = 45, hjust = 1, vjust = 1, size = 9),
    axis.text.y  = element_text(size = 9),
    plot.title   = element_text(hjust = 0.5, face = "bold")
  ) +
  labs(title = "Spearman Correlation – Cscore (all samples)", x = "", y = "")
print(spear_p)
ggsave(file.path(plots_dir, "spearman_corr.pdf"), spear_p, width = 10, height = 6)
```

## 3.3 Barplots

```{r QC barplots}
# ============================================================================
# QC barplot — total number of bins per sample
# ============================================================================

sample_comp <- bind_rows(
  data.frame(Sample    = "REH",
             Condition = "REH",
             Total     = nrow(corrected$REH)),
  data.frame(Sample    = "EP1_1",
             Condition = "EP1",
             Total     = nrow(corrected$EP1_1)),
  data.frame(Sample    = "EP1_2",
             Condition = "EP1",
             Total     = nrow(corrected$EP1_2)),
  data.frame(Sample    = "EP1_merge",
             Condition = "EP1",
             Total     = nrow(corrected$EP1_merge)),
  data.frame(Sample    = "EP1_Aro_1",
             Condition = "EP1_Aro",
             Total     = nrow(corrected$EP1_Aro_1)),
  data.frame(Sample    = "EP1_Aro_2",
             Condition = "EP1_Aro",
             Total     = nrow(corrected$EP1_Aro_2)),
  data.frame(Sample    = "EP1_Aro_merge",
             Condition = "EP1_Aro",
             Total     = nrow(corrected$EP1_Aro_merge))
) %>%
  mutate(Sample = factor(Sample, levels = c("REH",
                                             "EP1_1", "EP1_2", "EP1_merge",
                                             "EP1_Aro_1", "EP1_Aro_2", "EP1_Aro_merge")))

bins_QC <- ggplot(sample_comp, aes(x = Sample, y = Total, fill = Condition)) +
  geom_bar(stat      = "identity",
           color     = "black",
           linewidth = 0.35,
           width     = 0.75) +
  geom_text(aes(label = scales::comma(Total)),
            vjust    = -0.4,
            size     = 3.5,
            fontface = "bold") +
  scale_fill_manual(values = cond_colors, name = "Condition") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)),
                     labels = scales::comma) +
  theme_bw() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank(),
    axis.text.x        = element_text(size = 11, angle = 35,
                                      hjust = 1, face = "bold"),
    axis.text.y        = element_text(size = 10),
    axis.title         = element_text(size = 13),
    plot.title         = element_text(hjust = 0.5, face = "bold", size = 14),
    legend.title       = element_text(face = "bold")
  ) +
  xlab("") + ylab("Number of bins") +
  labs(title = "Total bins per sample (QC)")
print(bins_QC)
ggsave(file.path(plots_dir, "bins_x_sample_QC.pdf"), bins_QC, width = 10, height = 6)
```
---

# 4. A/B Compartment classification

> **All downstream analyses use the corrected merged data (`cs_REH_corr`).**

Bins with **cscore > 0** are assigned to **compartment A** (open, active chromatin)
and bins with **cscore ≤ 0** to **compartment B** (closed, inactive chromatin).
Classification is performed on the **merged** samples only for downstream
biological interpretation.

```{r compartment-classification}
cs_REH_merge <- cs_REH_corr %>%
  select(chr, start, end,
         score_REH, score_EP1_merge, score_EP1_Aro_merge) %>%
  mutate(
    comp_REH     = ifelse(score_REH > 0,             "A", "B"),
    comp_EP1     = ifelse(score_EP1_merge > 0,       "A", "B"),
    comp_EP1_Aro = ifelse(score_EP1_Aro_merge > 0,   "A", "B")
  )

cs_REH_merge <- cs_REH_merge %>%
  mutate(
    dyn_REH_EP1    = paste(comp_REH, comp_EP1,     sep = "_"),
    dyn_EP1_EP1Aro = paste(comp_EP1, comp_EP1_Aro, sep = "_"),
    dyn_all        = paste(comp_REH, comp_EP1, comp_EP1_Aro, sep = "_")
  )

# Summary table
summary_comp <- data.frame(
  Condition = c("REH", "EP1", "EP1_Aro"),
  A_bins    = c(sum(cs_REH_merge$comp_REH == "A"),
                sum(cs_REH_merge$comp_EP1 == "A"),
                sum(cs_REH_merge$comp_EP1_Aro == "A")),
  B_bins    = c(sum(cs_REH_merge$comp_REH == "B"),
                sum(cs_REH_merge$comp_EP1 == "B"),
                sum(cs_REH_merge$comp_EP1_Aro == "B"))
) %>%
  mutate(
    Total   = A_bins + B_bins,
    Pct_A   = round(A_bins / Total * 100, 1),
    Pct_B   = round(B_bins / Total * 100, 1)
  )

knitr::kable(summary_comp, caption = "A/B compartment counts per condition (merged samples)")
write.table(summary_comp,
            file.path(files_dir, "summary_comp.tsv"),
            sep = "\t", row.names = FALSE, quote = FALSE)
```

## 4.1 A/B barplot

```{r ab-barplot}
dt_merge <- rbind(
  table(cs_REH_merge$comp_REH),
  table(cs_REH_merge$comp_EP1),
  table(cs_REH_merge$comp_EP1_Aro)
)
rownames(dt_merge) <- c("REH", "EP1", "EP1_Aro")

AB_barp <- ggplot(melt(dt_merge), aes(fill = Var1, x = Var2, y = value)) +
  geom_bar(position = "dodge", stat = "identity",
           width = 0.7, color = "black") +
  scale_fill_manual(values = c("REH" = "#e74c3c",
                               "EP1" = "#3498db",
                               "EP1_Aro" = "#2ecc71")) +
  theme_bw() +
  theme(
    panel.grid   = element_blank(),
    axis.text    = element_text(size = 12, face = "bold"),
    axis.title   = element_text(size = 14),
    plot.title   = element_text(size = 15, face = "bold", hjust = 0.5),
    legend.title = element_blank()
  ) +
  xlab("") + ylab("Number of bins") +
  labs(title = "A/B compartments – merged samples")
print(AB_barp)
ggsave(file.path(plots_dir, "AB_counts.pdf"), AB_barp, width = 10, height = 6)
```

```{r Proportion barplot}
# Build proportion data from all samples (replicates + merges)
prop_df <- data.frame(
  Sample = c("REH", "EP1_1", "EP1_2", "EP1_Aro_1", "EP1_Aro_2", "EP1_merge", "EP1_Aro_merge"),
  A = c(sum(cs_REH_merge$comp_REH == "A"),
        sum(corrected$EP1_1$score_EP1_1 > 0),
        sum(corrected$EP1_2$score_EP1_2 > 0),
        sum(corrected$EP1_Aro_1$score_EP1_Aro_1 > 0),
        sum(corrected$EP1_Aro_2$score_EP1_Aro_2 > 0),
        sum(cs_REH_merge$comp_EP1 == "A"),
        sum(cs_REH_merge$comp_EP1_Aro == "A")),
  B = c(sum(cs_REH_merge$comp_REH == "B"),
        sum(corrected$EP1_1$score_EP1_1 <= 0),
        sum(corrected$EP1_2$score_EP1_2 <= 0),
        sum(corrected$EP1_Aro_1$score_EP1_Aro_1 <= 0),
        sum(corrected$EP1_Aro_2$score_EP1_Aro_2 <= 0),
        sum(cs_REH_merge$comp_EP1 == "B"),
        sum(cs_REH_merge$comp_EP1_Aro == "B"))
) %>%
  mutate(Total  = A + B,
         Pct_A  = A / Total * 100,
         Pct_B  = B / Total * 100) %>%
  pivot_longer(cols = c(Pct_A, Pct_B),
               names_to  = "Compartment",
               values_to = "Proportion") %>%
  mutate(Compartment = recode(Compartment, "Pct_A" = "A", "Pct_B" = "B"),
         Sample = factor(Sample, levels = c("REH","EP1_1","EP1_2",
                                             "EP1_merge","EP1_Aro_1",
                                             "EP1_Aro_2","EP1_Aro_merge")))

prop_bar <- ggplot(prop_df, aes(x = Sample, y = Proportion, fill = Compartment)) +
  geom_bar(stat = "identity", color = "black", linewidth = 0.4) +
  geom_text(aes(label = paste0(round(Proportion, 1), "%")),
            position = position_stack(vjust = 0.5),
            size = 3.8, fontface = "bold", color = "white") +
  scale_fill_manual(values = c("A" = "#f0a500", "B" = "#6c3483")) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 101)) +
  theme_bw() +
  theme(
    panel.grid   = element_blank(),
    axis.text.x  = element_text(size = 11, angle = 30, hjust = 1, face = "bold"),
    axis.text.y  = element_text(size = 10),
    axis.title   = element_text(size = 13),
    plot.title   = element_text(size = 14, face = "bold", hjust = 0.5),
    legend.title = element_text(face = "bold")
  ) +
  xlab("") + ylab("Proportion (%)") +
  labs(title = "A/B compartment proportions per sample")
print(prop_bar)
ggsave(file.path(plots_dir, "AB_proportions.pdf"), prop_bar, width = 10, height = 6)
```

---

# 5. Compartment switching patterns across experimental conditions

Although the conditions are shown in the order REH → EP1 → EP1_Aro for visualization purposes, these samples represent independent transfections rather than a true temporal series. Therefore, the switching classes describe compartment state combinations across conditions, not a biological time course.

Each genomic bin is tracked across all three conditions and assigned one of
**8 switching classes** (2³ combinations of A/B states, named as REH_EP1_EP1Aro).

| Switching | Interpretation |
|-----------|----------------|
| A_A_A | Constitutively active |
| B_B_B | Constitutively inactive |
| B_A_A | Activation at EP1 variants |
| B_B_A | Aro-specific activation |
| A_B_B | Inactivation at EP1 variants |
| A_A_B | Aro-specific inactivation |
| A_B_A | EP1-specific inactivation |
| B_A_B | EP1-specific activation |

## 5.1 Switching classification

```{r switching-classification}
# Define switching levels and colors keyed directly on A/B state codes
switching_levels <- c(
  "A_A_A", "B_B_B",   # constitutive
  "B_A_A", "B_B_A",   # gained A
  "A_B_B", "A_A_B",   # gained B
  "A_B_A", "B_A_B"    # transient
)

switching_colors <- c(
  "A_A_A" = "#2ecc71",   # green        – constitutive A
  "B_B_B" = "#3498db",   # blue         – constitutive B
  "B_A_A" = "#e67e22",   # orange       – B→A at EP1, maintained
  "B_B_A" = "#f1c40f",   # yellow       – B→A only at EP1_Aro
  "A_B_B" = "#8e44ad",   # purple       – A→B at EP1, maintained
  "A_A_B" = "#e74c3c",   # red          – A→B only at EP1_Aro
  "A_B_A" = "#c0392b",   # dark red     – A→B→A (EP1-specific B)
  "B_A_B" = "#1abc9c"    # teal         – B→A→B (EP1-specific A)
)

cs_REH_merge$switching <- factor(cs_REH_merge$dyn_all,
                                   levels = switching_levels)

switch_counts <- as.data.frame(table(switching = cs_REH_merge$switching)) %>%
  mutate(
    Pct   = round(Freq / nrow(cs_REH_merge) * 100, 2),
    Group = case_when(
      switching %in% c("A_A_A", "B_B_B") ~ "Constitutive",
      switching %in% c("B_A_A", "B_B_A") ~ "Gain A",
      switching %in% c("A_B_B", "A_A_B") ~ "Gain B",
      TRUE                                 ~ "Transient"
    )
  )

knitr::kable(switch_counts[, c("switching", "Freq", "Pct", "Group")],
             col.names = c("Switching (REH_EP1_EP1Aro)", "Bins", "% total", "Group"),
             caption   = "Bin counts per switching class across REH → EP1 → EP1_Aro")
write.table(switch_counts,
            file.path(files_dir, "switch_counts.tsv"),
            sep = "\t", row.names = FALSE, quote = FALSE)
```

## 5.2 Bar plot of switching counts

```{r switching-barplot, fig.width=10, fig.height=5}
# Exclude constitutive bins for a clearer view of dynamics
switch_dynamic <- switch_counts[!switch_counts$switching %in% c("A_A_A", "B_B_B"), ]

switch_bar <- ggplot(switch_dynamic,
       aes(x    = reorder(switching, -Freq),
           y    = Freq,
           fill = switching)) +
  geom_bar(stat = "identity", color = "black", width = 0.7) +
  geom_text(aes(label = paste0(Pct, "%")),
            vjust = -0.4, size = 3.5, fontface = "bold") +
  scale_fill_manual(values = switching_colors) +
  theme_bw() +
  theme(
    panel.grid      = element_blank(),
    axis.text.x     = element_text(size = 11, angle = 30, hjust = 1, face = "bold"),
    axis.text.y     = element_text(size = 11),
    axis.title      = element_text(size = 13),
    plot.title      = element_text(size = 15, face = "bold", hjust = 0.5),
    legend.position = "none"
  ) +
  xlab("Switching class (REH_EP1_EP1Aro)") + ylab("Number of bins") +
  labs(title = "Compartment switching patterns across conditions (REH, EP1, EP1_Aro)\n(constitutive bins excluded)")
print(switch_bar)
ggsave(file.path(plots_dir, "switching_barplot.pdf"), switch_bar, width = 10, height = 6)
```

## 5.3 Alluvial / Sankey plot of transitions

The alluvial plot traces each bin's compartment state at each condition,
making it easy to see the dominant flows between states.

```{r Alluvial / Sankey plot}
library(ggalluvial)

alluv_sum <- cs_REH_merge %>%
  select(switching,
         REH     = comp_REH,
         EP1     = comp_EP1,
         EP1_Aro = comp_EP1_Aro) %>%
  pivot_longer(cols      = c(REH, EP1, EP1_Aro),
               names_to  = "Condition",
               values_to = "Compartment") %>%
  mutate(
    Condition = factor(Condition, levels = c("REH", "EP1", "EP1_Aro")),
    switching = factor(switching, levels = switching_levels)
  ) %>%
  group_by(switching, Condition, Compartment) %>%
  summarise(Freq = n(), .groups = "drop")

alluv_p <- ggplot(alluv_sum,
       aes(x        = Condition,
           y        = Freq,
           stratum  = Compartment,
           alluvium = switching,
           fill     = switching,
           label    = Compartment)) +
  geom_alluvium(alpha     = 0.55,
                color     = NA,
                linewidth = 0) +
  geom_stratum(alpha     = 0.88,
               color     = "white",
               linewidth = 0.6) +
  geom_text(stat     = "stratum",
            size     = 5,
            fontface = "bold",
            color    = "white") +
  scale_fill_manual(values = switching_colors,
                    name   = "Switching\n(REH_EP1_EP1Aro)") +
  scale_y_continuous(expand = c(0, 0)) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid      = element_blank(),
    axis.text.x     = element_text(size = 14, face = "bold", color = "gray20"),
    axis.text.y     = element_text(size = 10, color = "gray40"),
    axis.title.y    = element_text(size = 13, color = "gray30"),
    plot.title      = element_text(size = 15, face = "bold", hjust = 0.5,
                                   margin = margin(b = 12)),
    plot.background = element_rect(fill = "gray97", color = NA),
    legend.text     = element_text(size = 9),
    legend.title    = element_text(size = 10, face = "bold")
  ) +
  xlab("") + ylab("Number of bins") +
  labs(title = "Compartment state distribution across conditions (REH, EP1, EP1_Aro)")
print(alluv_p)
ggsave(file.path(plots_dir, "switch_alluvian.pdf"), alluv_p, width = 10, height = 6)
```

## 5.4 Pie chart of dynamic bins only

Among bins that change state at least once across the three conditions.

```{r pie-dynamic, fig.width=7, fig.height=7}
dynamic_bins <- cs_REH_merge %>%
  filter(!switching %in% c("A_A_A", "B_B_B"))

n_dynamic <- nrow(dynamic_bins)
n_total   <- nrow(cs_REH_merge)

pie_df <- as.data.frame(table(switching = dynamic_bins$switching)) %>%
  filter(Freq > 0) %>%
  mutate(
    Pct      = round(Freq / n_dynamic * 100, 1),
    PieLabel = paste0(Pct, "%")
  )

switch_pie <- ggplot(pie_df, aes(x = "", y = Freq, fill = switching)) +
  geom_col(width = 1, color = "black", linewidth = 1.1) +
  geom_text(aes(label = paste0(switching, "\n", PieLabel)),
            position = position_stack(vjust = 0.5),
            size = 3.5, fontface = "bold", color = "white") +
  coord_polar(theta = "y", start = 0) +
  scale_fill_manual(values = switching_colors,
                    name   = "Switching\n(REH_EP1_EP1Aro)") +
  theme_void() +
  ggtitle(paste0("Dynamic bins: n = ", n_dynamic,
                 " (", round(n_dynamic / n_total * 100, 1), "% of all bins)")) +
  theme(
    plot.title      = element_text(hjust = 0.5, size = 13, face = "bold"),
    legend.text     = element_text(size = 10),
    legend.title    = element_text(size = 11, face = "bold"),
    legend.position = "right"
  )
print(switch_pie)
ggsave(file.path(plots_dir, "switching_pie.pdf"), switch_pie, width = 10, height = 6)
```

## 5.5 Summary: switching groups

```{r switching-summary-table}
group_summary <- switch_counts %>%
  group_by(Group) %>%
  summarise(
    Bins             = sum(Freq),
    Pct_total        = round(sum(Freq) / nrow(cs_REH_merge) * 100, 1),
    Switching_classes = paste(switching, collapse = ", ")
  ) %>%
  arrange(desc(Bins))

knitr::kable(group_summary,
             col.names = c("Group", "Bins", "% of total", "Switching classes"),
             caption   = "Compartment switching groups – REH → EP1 → EP1_Aro")
write.table(group_summary,
            file.path(files_dir, "switch_groups.tsv"),
            sep = "\t", row.names = FALSE, quote = FALSE)
```

## 5.6 Heatmap

```{r heatmap, fig.width=12, fig.height=16, fig.align="center"}
# Filter dynamic bins and complete cases
dynamic_bins_complete <- cs_REH_merge %>%
  filter(!switching %in% c("A_A_A", "B_B_B")) %>%
  filter(!is.na(score_REH) & !is.na(score_EP1_merge) & !is.na(score_EP1_Aro_merge)) %>%
  arrange(switching)

# Prepare heatmap matrix
cscore_heat_dyn <- as.matrix(dynamic_bins_complete[, c("score_REH", "score_EP1_merge", "score_EP1_Aro_merge")])
rownames(cscore_heat_dyn) <- rownames(dynamic_bins_complete)
colnames(cscore_heat_dyn) <- c("REH", "EP1", "EP1_Aro")

# Prepare row annotation using switching codes directly
annot_row <- data.frame(
  Switching = factor(dynamic_bins_complete$switching,
                     levels = switching_levels[!switching_levels %in% c("A_A_A", "B_B_B")])
)
rownames(annot_row) <- rownames(cscore_heat_dyn)

# Annotation colors keyed on switching codes
annot_colors <- list(
  Switching = switching_colors[levels(annot_row$Switching)]
)

# Heatmap colors
heat_colors <- colorRampPalette(c("#2166AC", "white", "#B2182B"))(1000)

# Plot heatmap
pdf(file.path(plots_dir, "cscore_heatmap.pdf"), width = 10, height = 14)
pheatmap(cscore_heat_dyn,
         color                = heat_colors,
         cluster_rows         = FALSE,
         cluster_cols         = FALSE,
         show_rownames        = FALSE,
         annotation_row       = annot_row,
         annotation_colors    = annot_colors,
         annotation_names_row = FALSE,
         treeheight_row       = 50,
         scale                = "row",
         fontsize             = 12,
         fontsize_col         = 13,
         angle_col            = 45,
         legend_breaks        = c(-1, -0.5, 0, 0.5, 1),
         main                 = "Cscore Heatmap – Dynamic bins only (REH / EP1 / EP1_Aro)",
         mar                  = c(4, 6, 4, 10))
dev.off()

pheatmap(cscore_heat_dyn,
         color                = heat_colors,
         cluster_rows         = FALSE,
         cluster_cols         = FALSE,
         show_rownames        = FALSE,
         annotation_row       = annot_row,
         annotation_colors    = annot_colors,
         annotation_names_row = FALSE,
         treeheight_row       = 50,
         scale                = "row",
         fontsize             = 12,
         fontsize_col         = 13,
         angle_col            = 45,
         legend_breaks        = c(-1, -0.5, 0, 0.5, 1),
         main                 = "Cscore Heatmap – Dynamic bins only (REH / EP1 / EP1_Aro)",
         mar                  = c(4, 6, 4, 10))

```

---

# 6. Gene enrichment analysis on dynamic compartments

```{r Enrichment analysis}
# ============================================================================
# Compartment Switching – Enrichment Analysis (GenomicRanges + enrichR)
# ============================================================================
library(GenomicRanges)
library(enrichR)
library(dplyr)
library(ggplot2)

# ----------------------------------------------------------------------------
# 1. Load gene annotation BED (gencode v28)
# ----------------------------------------------------------------------------
genes_bed <- read.table("R:/R_STIK/xavi_scripts/genes_gencode.v28.bed",
                        sep       = "\t",
                        header    = FALSE,
                        col.names = c("chrom", "start", "end", "strand",
                                      "gene_name", "ID", "description")) %>%
  select(chrom, start, end, gene_name) %>%
  mutate(start = as.integer(start),
         end   = as.integer(end))

genes_gr <- makeGRangesFromDataFrame(genes_bed,
                                     seqnames.field     = "chrom",
                                     start.field        = "start",
                                     end.field          = "end",
                                     keep.extra.columns = TRUE)

# ----------------------------------------------------------------------------
# 2. Helper: get gene symbols overlapping a set of bins
# ----------------------------------------------------------------------------
get_genes_gr <- function(sw_class) {

  bins <- cs_REH_merge %>%
    filter(switching == sw_class) %>%
    select(chrom = chr, start, end) %>%
    mutate(start = as.integer(start),
           end   = as.integer(end))

  if (nrow(bins) == 0) return(character(0))

  bins_gr <- makeGRangesFromDataFrame(bins,
                                      seqnames.field = "chrom",
                                      start.field    = "start",
                                      end.field      = "end")

  hits <- findOverlaps(bins_gr, genes_gr)
  unique(genes_gr$gene_name[subjectHits(hits)])
}

# ----------------------------------------------------------------------------
# 3. Build gene list for all 6 dynamic switching classes
# ----------------------------------------------------------------------------
switching_dynamic_levels <- c("B_A_A", "B_B_A", "A_B_B", "A_A_B", "A_B_A", "B_A_B")

class_counts <- table(cs_REH_merge$switching)
cat("\nBin counts per dynamic switching class:\n")
print(class_counts[switching_dynamic_levels])

missing <- switching_dynamic_levels[
  !switching_dynamic_levels %in% names(class_counts) |
   class_counts[switching_dynamic_levels] == 0
]
if (length(missing) > 0) {
  cat("\nWARNING: These classes have 0 bins and will be skipped:", paste(missing, collapse = ", "), "\n")
} else {
  cat("\nAll 6 dynamic classes present in data. Proceeding with gene overlap.\n")
}

switching_gene_lists <- lapply(
  setNames(switching_dynamic_levels, switching_dynamic_levels),
  get_genes_gr
)

cat("\nGenes overlapping each switching class:\n")
gene_summary <- data.frame(
  switching = switching_dynamic_levels,
  n_genes   = sapply(switching_gene_lists, length)
)
print(gene_summary)

too_few <- gene_summary[gene_summary$n_genes < 5, ]
if (nrow(too_few) > 0) {
  cat("\nWARNING: The following classes have <5 genes and will be skipped by enrichR:\n")
  print(too_few)
}

# ----------------------------------------------------------------------------
# 4. Databases
# ----------------------------------------------------------------------------
dbs <- c("GO_Biological_Process_2023",
         "KEGG_2021_Human",
         "WikiPathway_2023_Human",
         "Reactome_2022",
         "MSigDB_Hallmark_2020")

websiteLive <- getOption("enrichR.live", TRUE)
setEnrichrSite("Enrichr")

# ----------------------------------------------------------------------------
# 5. Helpers
# ----------------------------------------------------------------------------
process_enrichr_db <- function(db_result) {
  df <- db_result %>%
    separate(col = "Overlap", into = c("Gene_count", "total"), sep = "/",
             convert = TRUE, extra = "drop", fill = "right")
  df$gene_ratio <- df$Gene_count / df$total
  df$Term <- gsub("\\s*\\(GO:\\d+\\)", "",  df$Term)
  df$Term <- gsub("\\s*R-HSA-[0-9.]+",  "", df$Term)
  df$Term <- gsub("\\s*WP\\d+",         "", df$Term)
  df$Term <- trimws(df$Term)
  df
}

save_enrichr_genelists <- function(sig_terms, prefix, out_dir) {
  sig_out <- sig_terms %>%
    select(Term, Gene_count, total, gene_ratio, P.value, Adjusted.P.value, Genes) %>%
    arrange(Adjusted.P.value)
  write.table(sig_out,
              file.path(out_dir, paste0(prefix, "_enrichment.tsv")),
              sep = "\t", row.names = FALSE, quote = FALSE)
}

plot_enrichr <- function(sig_terms, title_str, xlab_str) {
  plot_data <- head(sig_terms, 20)
  ggplot(plot_data, aes(x = Term, y = -log10(Adjusted.P.value))) +
    geom_hline(yintercept = 1, linetype = "dashed", color = "azure4", size = 0.5) +
    geom_point(aes(size = Gene_count, colour = gene_ratio), alpha = 0.7) +
    scale_x_discrete(limits = rev(plot_data$Term)) +
    scale_color_gradient(low = "green", high = "red", limits = c(0, NA)) +
    coord_flip() +
    theme_bw() +
    theme(axis.ticks.length  = unit(-0.1, "cm"),
          axis.text.x        = element_text(margin = margin(5, 5, 0, 5, "pt")),
          axis.text.y        = element_text(margin = margin(5, 5, 5, 5, "pt")),
          axis.text          = element_text(color = "black"),
          panel.grid.minor   = element_blank(),
          legend.title.align = 0.5,
          plot.title         = element_text(hjust = 0.5, color = "black",
                                            face = "bold", size = 14)) +
    xlab(xlab_str) +
    ylab("-log10(Adjusted P-value)") +
    labs(color = "Gene Ratio", size = "Number\nof genes") +
    ggtitle(title_str)
}

# ----------------------------------------------------------------------------
# 6. Main enrichR loop
# ----------------------------------------------------------------------------
enrichr_compartment_results <- list()

for (sw_class in switching_dynamic_levels) {

  gene_vec <- switching_gene_lists[[sw_class]]

  cat("\n=== Switching:", sw_class,
      "| Genes:", length(gene_vec), "===\n")

  if (length(gene_vec) < 5 || !websiteLive) {
    cat("  Skipping – too few genes or enrichR offline\n")
    next
  }

  enriched <- enrichr(gene_vec, dbs)
  enrichr_compartment_results[[sw_class]] <- enriched

  for (db_idx in seq_along(dbs)) {
    db_name <- dbs[db_idx]

    if (is.null(enriched[[db_idx]]) || nrow(enriched[[db_idx]]) == 0) next

    enriched_processed <- process_enrichr_db(enriched[[db_idx]])
    sig_terms          <- subset(enriched_processed, Adjusted.P.value < 0.05)

    if (nrow(sig_terms) == 0) {
      cat("  No significant terms:", db_name, "\n")
      next
    }

    save_enrichr_genelists(
      sig_terms,
      prefix  = paste0(sw_class, "_", db_name),
      out_dir = files_dir
    )

    p <- plot_enrichr(
      sig_terms,
      title_str = paste0(db_name, "\nSwitching: ", sw_class),
      xlab_str  = paste0(db_name, " Terms")
    )
    print(p)

    ggsave(
      filename = file.path(plots_dir,
                           paste0("enrichR_", db_name, "_", sw_class, ".pdf")),
      plot   = p,
      width  = 10,
      height = 7
    )

    cat("  Saved:", db_name, "\n")
  }
}

cat("\n=== Compartment enrichR analysis complete ===\n")
```

---

# 7. ChIP-Seq EP1 intersection

```{r ChIP-Seq intersection}
# ============================================================================
# Intersection of switching compartments with ChIP-seq peaks
# ============================================================================
library(GenomicRanges)
library(ggplot2)
library(dplyr)
library(tidyr)

# ----------------------------------------------------------------------------
# Load ChIP-seq peaks BED
# ----------------------------------------------------------------------------
peaks_bed <- read.table("R:/R_STIK/chip_seq/Xavi_TFM/697HF_E2APBX1-HF_peaks.bed",
                        sep       = "\t",
                        header    = FALSE,
                        col.names = c("chrom", "start", "end")
                        [1:ncol(read.table(
                        "R:/R_STIK/chip_seq/Xavi_TFM/697HF_E2APBX1-HF_peaks.bed",
                        sep = "\t", nrows = 1))]) %>%
  select(chrom = 1, start = 2, end = 3) %>%
  mutate(start = as.integer(start),
         end   = as.integer(end))

peaks_gr <- makeGRangesFromDataFrame(peaks_bed,
                                     seqnames.field = "chrom",
                                     start.field    = "start",
                                     end.field      = "end")

cat("Total peaks loaded:", length(peaks_gr), "\n")

# ----------------------------------------------------------------------------
# Overlap peaks with every switching class (+ constitutive as background)
# ----------------------------------------------------------------------------
all_switching_levels <- c("A_A_A", "B_B_B",
                           "B_A_A", "B_B_A",
                           "A_B_B", "A_A_B",
                           "A_B_A", "B_A_B")

overlap_counts <- lapply(all_switching_levels, function(sw) {

  bins <- cs_REH_merge %>%
    filter(switching == sw) %>%
    select(chrom = chr, start, end) %>%
    mutate(start = as.integer(start),
           end   = as.integer(end))

  bins_gr <- makeGRangesFromDataFrame(bins,
                                      seqnames.field = "chrom",
                                      start.field    = "start",
                                      end.field      = "end")

  n_bins      <- length(bins_gr)
  n_overlap   <- length(unique(queryHits(findOverlaps(bins_gr, peaks_gr))))
  pct_overlap <- n_overlap / n_bins * 100

  data.frame(switching   = sw,
             n_bins      = n_bins,
             n_overlap   = n_overlap,
             pct_overlap = pct_overlap)
}) %>% bind_rows()

# Compute fold enrichment over genome-wide background
background_pct <- sum(overlap_counts$n_overlap) / sum(overlap_counts$n_bins) * 100
overlap_counts <- overlap_counts %>%
  mutate(
    fold_enrichment = pct_overlap / background_pct,
    switching       = factor(switching, levels = all_switching_levels)
  )

print(overlap_counts[, c("switching", "n_bins", "n_overlap",
                          "pct_overlap", "fold_enrichment")])

# ----------------------------------------------------------------------------
# Annotate each bin with peak overlap for boxplot
# ----------------------------------------------------------------------------
all_bins_gr <- makeGRangesFromDataFrame(
  cs_REH_merge %>% select(chr, start, end),
  seqnames.field = "chr",
  start.field    = "start",
  end.field      = "end"
)

cs_REH_merge <- cs_REH_merge %>%
  mutate(
    peak_overlap = ifelse(countOverlaps(all_bins_gr, peaks_gr) > 0, "Peak", "No peak")
  )

# ----------------------------------------------------------------------------
# Plot 1 — Polar enrichment plot with all 8 switching classes
# ----------------------------------------------------------------------------
overlap_polar <- ggplot(overlap_counts,
                        aes(x = switching, y = fold_enrichment, fill = switching)) +
  geom_col(color = "black", linewidth = 0.4, width = 0.85) +
  geom_hline(yintercept = 1, linetype = "dashed",
             color = "gray40", linewidth = 0.5) +
  coord_polar() +
  scale_fill_manual(values = switching_colors,
                    name   = "Switching\n(REH_EP1_EP1Aro)") +
  scale_y_continuous(limits = c(0, max(overlap_counts$fold_enrichment) * 1.2)) +
  theme_bw() +
  theme(
    axis.text.x  = element_text(size = 8, face = "bold"),
    axis.text.y  = element_text(size = 8),
    axis.title.x = element_blank(),
    axis.title.y = element_text(size = 11),
    plot.title   = element_text(hjust = 0.5, face = "bold", size = 13),
    panel.grid   = element_line(color = "gray85", linetype = "dashed"),
    legend.title = element_text(face = "bold")
  ) +
  ylab("Fold Enrichment over background") +
  labs(title = "E2A-PBX1 peak enrichment per compartment switching class\n(REH_EP1_EP1Aro)")
print(overlap_polar)
ggsave(file.path(plots_dir, "EP1_overlap_polar.pdf"), overlap_polar, width = 10, height = 6)

# ----------------------------------------------------------------------------
# Boxplot — cscore per condition, faceted by peak overlap
# ----------------------------------------------------------------------------
box_df <- cs_REH_merge %>%
  select(peak_overlap,
         REH     = score_REH,
         EP1     = score_EP1_merge,
         EP1_Aro = score_EP1_Aro_merge) %>%
  pivot_longer(cols      = c(REH, EP1, EP1_Aro),
               names_to  = "Condition",
               values_to = "cscore") %>%
  mutate(
    Condition    = factor(Condition, levels = c("REH", "EP1", "EP1_Aro")),
    peak_overlap = factor(peak_overlap, levels = c("Peak", "No peak")),
    fill_group   = ifelse(peak_overlap == "Peak", as.character(Condition), "No peak")
  )

fill_colors <- c(cond_colors, "No peak" = "#aaaaaa")

cscore_EP1_box <- ggplot(box_df, aes(x = Condition, y = cscore, fill = fill_group)) +
  geom_boxplot(outlier.size  = 0.3,
               outlier.alpha = 0.3,
               linewidth     = 0.45,
               alpha         = 0.9) +
  geom_hline(yintercept = 0, linetype = "dashed",
             color = "gray30", linewidth = 0.4) +
  facet_wrap(~ peak_overlap, nrow = 1) +
  scale_fill_manual(values = fill_colors,
                    breaks = names(cond_colors),
                    name   = "Condition") +
  scale_y_continuous(limits = c(-1, 1),
                     breaks = c(-1, -0.5, 0, 0.5, 1)) +
  theme_bw() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank(),
    axis.text.x        = element_text(size = 11, angle = 30,
                                      hjust = 1, face = "bold"),
    axis.text.y        = element_text(size = 10),
    axis.title         = element_text(size = 13),
    strip.text         = element_text(size = 12, face = "bold"),
    strip.background   = element_rect(fill = "gray92", color = "black"),
    plot.title         = element_text(hjust = 0.5, face = "bold", size = 13),
    legend.title       = element_text(face = "bold", size = 10),
    legend.text        = element_text(size = 9)
  ) +
  xlab("Condition") + ylab("cscore") +
  labs(title = "cscore distribution – bins overlapping E2A-PBX1 peaks vs non-overlapping")
print(cscore_EP1_box)
ggsave(file.path(plots_dir, "EP1_overlap_cscore.pdf"), cscore_EP1_box, width = 10, height = 6)
```

---

# 8. Save outputs

```{r save}
write.table(cs_REH_merge,
            file.path(files_dir, "cs_REH_merge.tab"),
            sep       = "\t",
            quote     = FALSE,
            col.names = TRUE,
            row.names = FALSE)

cat("Merged table saved to files/cs_REH_merge.tab\n")
```

---

# Session info

```{r session-info}
sessionInfo()
```
