---
title: "Hi-C Compartment and Loop Analysis – REH samples"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: hide
    fig_width: 8
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)
knitr::opts_knit$set(root.dir = "R:/R_STIK/Hi-C/HiC_marta_Oct25/cscore")
# knitr::opts_knit$set(root.dir = "/ijc/LABS/STIK/RAW/Hi-C/HiC_marta_Oct25/cscore")

```

---

# Overview

This notebook analyses **A/B chromatin compartments** inferred from Hi-C data using
Cscore bedgraph files for three conditions:

| Condition | Description | Files |
|-----------|-------------|-------|
| **REH** | Parental cell line | `REH_tp.bedgraph_correct.bedgraph` |
| **EP1** | EP1 treatment | `HiC-REH-EP1-{1,2}_tp…` + merged |
| **EP1-Aro** | EP1 + Aromatic treatment | `HiC-REH-EP1-Aro-{1,2}_tp…` + merged |

**Analysis steps:**

1. Load & merge all bedgraph files  
2. QC: PCA and Spearman correlation across replicates and merged samples  
3. A/B compartment classification  
4. Compartment switching dynamics  
5. Heatmap of cscore values  

---

# 1. Libraries and output directories

```{r libraries}
library(dplyr)
library(ggplot2)
library(reshape2)
library(ggrepel)
library(pheatmap)
library(viridis)
library(tidyr)
library(grid)
library(cowplot)

output_dir <- "Comp_Analysis"

files_dir  <- file.path(output_dir, "files")
plots_dir  <- file.path(output_dir, "plots")

for (d in c(files_dir, plots_dir)) {
  if (!dir.exists(d)) dir.create(d, recursive = TRUE)
}
```

---

# 2. Data loading

All bedgraph files are loaded from the cscore iteration trial directory.
Each file contains four columns: `chr`, `start`, `end`, and the cscore `score`.

```{r load-data}
data_dir <- "R:/R_STIK/Hi-C/HiC_marta_Oct25/cscore"
# data_dir <-"/ijc/LABS/STIK/RAW/Hi-C/HiC_marta_Oct25/cscore"

read_cscore <- function(filename, score_colname) {
  read.table(file.path(data_dir, filename),
             sep       = "\t",
             col.names = c("chr", "start", "end", score_colname))
}

# Individual replicates
REH        <- read_cscore("REH_tp.bedgraph_correct.bedgraph",           "score_REH")
EP1_1      <- read_cscore("HiC-REH-EP1-1_tp.bedgraph_correct.bedgraph", "score_EP1_1")
EP1_2      <- read_cscore("HiC-REH-EP1-2_tp.bedgraph_correct.bedgraph", "score_EP1_2")
EP1_Aro_1  <- read_cscore("HiC-REH-EP1-Aro-1_tp.bedgraph_correct.bedgraph", "score_EP1_Aro_1")
EP1_Aro_2  <- read_cscore("HiC-REH-EP1-Aro-2_tp.bedgraph_correct.bedgraph", "score_EP1_Aro_2")

# Merged replicates
EP1_merge     <- read_cscore("HiC-REH-EP1_tp.bedgraph_correct.bedgraph",     "score_EP1_merge")
EP1_Aro_merge <- read_cscore("HiC-REH-EP1-Aro_tp.bedgraph_correct.bedgraph", "score_EP1_Aro_merge")
```

## 2.1 Merge into a single dataframe

All files are merged by genomic coordinates (`chr`, `start`, `end`).

```{r merge-data}
cs_REH <- Reduce(
  function(x, y) merge(x, y, by = c("chr", "start", "end")),
  list(REH, EP1_1, EP1_2, EP1_Aro_1, EP1_Aro_2, EP1_merge, EP1_Aro_merge)
)

cat("Columns:", paste(colnames(cs_REH), collapse = ", "), "\n")

head(cs_REH)
```

---

# 3. Quality Control

## 3.1 PCA

PCA is run on the transposed cscore matrix (samples as rows, genomic bins as
columns) to assess whether:

- Replicates of the same condition cluster tightly together  
- Merged samples fall close to their corresponding replicates  
- The three conditions separate along the principal components  

**Shape legend:** circles = parental REH, squares = individual replicates, triangles = merged.  
**Colour legend:** condition.

```{r pca}
sample_labels <- c("REH", "EP1_1", "EP1_2", "EP1_Aro_1", "EP1_Aro_2",
                   "EP1_merge", "EP1_Aro_merge")

cscore_matrix           <- as.matrix(cs_REH[, 4:10])
colnames(cscore_matrix) <- sample_labels
cscore_matrix           <- na.omit(cscore_matrix)

# remove bins with zero variance
cscore_matrix <- cscore_matrix[apply(cscore_matrix, 1, var) != 0, ]

# PCA
PCA_REH <- prcomp(t(cscore_matrix), scale. = TRUE)
pct_var <- round(100 * PCA_REH$sdev^2 / sum(PCA_REH$sdev^2), 1)

# Build plot dataframe
pca_df            <- as.data.frame(PCA_REH$x)
pca_df$sample     <- sample_labels
pca_df$condition  <- c("REH", "EP1", "EP1", "EP1_Aro", "EP1_Aro", "EP1", "EP1_Aro")
pca_df$type       <- c("original", "replicate", "replicate",
                        "replicate", "replicate", "merge", "merge")

# Colour and shape scales
cond_colors <- c("REH"     = "#e74c3c",
                 "EP1"     = "#3498db",
                 "EP1_Aro" = "#2ecc71")

type_shapes <- c("original"  = 21,
                 "replicate" = 22,
                 "merge"     = 24)

pca_before <- ggplot(pca_df, aes(x = PC1, y = PC2,
                   fill  = condition,
                   shape = type,
                   label = sample)) +
  geom_point(color = "black", stroke = 1.3, size = 5) +
  geom_text_repel(size = 4, max.overlaps = 20) +
  scale_fill_manual(values = cond_colors) +
  scale_shape_manual(values = type_shapes) +
  theme_bw() +
  labs(
    title = "Cscore PCA – REH samples (replicates + merges)",
    x     = paste0("PC1 (", pct_var[1], "% variance)"),
    y     = paste0("PC2 (", pct_var[2], "% variance)")
  ) +
  theme(
    panel.grid   = element_blank(),
    plot.title   = element_text(size = 15, face = "bold", hjust = 0.5),
    axis.title   = element_text(size = 13),
    axis.text    = element_text(size = 11),
    legend.title = element_text(size = 11),
    legend.text  = element_text(size = 10)
  ) +
  guides(
    fill  = guide_legend(title = "Condition", override.aes = list(shape = 21)),
    shape = guide_legend(title = "Type")
  )
print(pca_before)
ggsave(file.path(plots_dir, "pca_before_correction.pdf"), pca_before, width = 10, height = 6)
```
Now we remove some problematic regions that we detected through IGV visualization of the bedgraphs

```{r Problematic regions removal + PCA after correction}
# Regions to remove
chr4_start  <- 127470171
chr4_end    <- 141890371

chr21_start <- 4514218
chr21_end   <- 12001870

# Function to correct a bedgraph
correct_bedgraph <- function(df, score_col, flip_chr13 = FALSE){

  # remove chr4 problematic region
  df <- df[!(df$chr == "chr4" &
             df$end   >= chr4_start &
             df$start <= chr4_end), ]

  # remove chr21 problematic region
  df <- df[!(df$chr == "chr21" &
             df$end   >= chr21_start &
             df$start <= chr21_end), ]

  # flip chr13 if required
  if(flip_chr13){
    df[df$chr == "chr13", score_col] <- -df[df$chr == "chr13", score_col]
  }

  return(df)
}

# List all samples
samples <- list(
  REH            = list(df = REH,           col = "score_REH",            flip = TRUE),
  EP1_1          = list(df = EP1_1,         col = "score_EP1_1",          flip = TRUE),
  EP1_2          = list(df = EP1_2,         col = "score_EP1_2",          flip = FALSE),
  EP1_Aro_1      = list(df = EP1_Aro_1,     col = "score_EP1_Aro_1",      flip = TRUE),
  EP1_Aro_2      = list(df = EP1_Aro_2,     col = "score_EP1_Aro_2",      flip = TRUE),
  EP1_merge      = list(df = EP1_merge,     col = "score_EP1_merge",      flip = TRUE),
  EP1_Aro_merge  = list(df = EP1_Aro_merge, col = "score_EP1_Aro_merge",  flip = TRUE)
)

# Apply correction to all samples
corrected <- list()

for(name in names(samples)){

  sample_info <- samples[[name]]

  df_corr <- correct_bedgraph(
    sample_info$df,
    sample_info$col,
    sample_info$flip
  )

  # store corrected dataframe
  corrected[[name]] <- df_corr

  outfile <- paste0(name, "_manual_corrected.bedgraph")
  
  write.table(
    df_corr,
    outfile,
    sep = "\t",
    quote = FALSE,
    row.names = FALSE,
    col.names = FALSE
  )

  cat("Saved:", outfile, "\n")
}
```

### Re-merge corrected samples

```{r merge-corrected}
cs_REH_corr <- Reduce(
  function(x, y) merge(x, y, by = c("chr", "start", "end")),
  list(corrected$REH,
       corrected$EP1_1,
       corrected$EP1_2,
       corrected$EP1_Aro_1,
       corrected$EP1_Aro_2,
       corrected$EP1_merge,
       corrected$EP1_Aro_merge)
)

cat("Bins in corrected merged dataframe:", nrow(cs_REH_corr), "\n")
```

## 3.4 PCA (post-correction)

Re-run the same PCA on the corrected data. Samples should now cluster more
cleanly by condition, with replicates tighter and the chr4/chr13 artefacts
no longer driving separation.

```{r pca-corrected}
cscore_matrix_corr           <- as.matrix(cs_REH_corr[, 4:10])
colnames(cscore_matrix_corr) <- sample_labels
cscore_matrix_corr           <- na.omit(cscore_matrix_corr)
cscore_matrix_corr           <- cscore_matrix_corr[apply(cscore_matrix_corr, 1, var) != 0, ]

cat("Bins used for post-correction PCA:", nrow(cscore_matrix_corr), "\n")

PCA_corr <- prcomp(t(cscore_matrix_corr), scale. = TRUE)
pct_var_corr <- round(100 * PCA_corr$sdev^2 / sum(PCA_corr$sdev^2), 1)

pca_df_corr            <- as.data.frame(PCA_corr$x)
pca_df_corr$sample     <- sample_labels
pca_df_corr$condition  <- c("REH", "EP1", "EP1", "EP1_Aro", "EP1_Aro", "EP1", "EP1_Aro")
pca_df_corr$type       <- c("original", "replicate", "replicate",
                             "replicate", "replicate", "merge", "merge")

pca_after <- ggplot(pca_df_corr, aes(x = PC1, y = PC2,
                        fill  = condition,
                        shape = type,
                        label = sample)) +
  geom_point(color = "black", stroke = 1.3, size = 5) +
  geom_text_repel(size = 4, max.overlaps = 20) +
  scale_fill_manual(values = cond_colors) +
  scale_shape_manual(values = type_shapes) +
  theme_bw() +
  labs(
    title = "Cscore PCA – post-correction (replicates + merges)",
    x     = paste0("PC1 (", pct_var_corr[1], "% variance)"),
    y     = paste0("PC2 (", pct_var_corr[2], "% variance)")
  ) +
  theme(
    panel.grid   = element_blank(),
    plot.title   = element_text(size = 15, face = "bold", hjust = 0.5),
    axis.title   = element_text(size = 13),
    axis.text    = element_text(size = 11),
    legend.title = element_text(size = 11),
    legend.text  = element_text(size = 10)
  ) +
  guides(
    fill  = guide_legend(title = "Condition", override.aes = list(shape = 21)),
    shape = guide_legend(title = "Type")
  )
print(pca_after)
ggsave(file.path(plots_dir, "pca_after_correction.pdf"), pca_after, width = 10, height = 6)
```

## 3.2 Spearman correlation matrix

Pairwise Spearman correlation between all samples. Values close to 1 between
replicates of the same condition indicate good reproducibility.

```{r correlation}
cor_mat <- cor(cscore_matrix, method = "spearman")

spear_p <- ggplot(melt(cor_mat), aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  geom_text(aes(label = round(value, 2)), size = 3.2) +
  coord_equal() +
  theme_bw() +
  theme(
    axis.text.x  = element_text(angle = 45, hjust = 1, vjust = 1, size = 9),
    axis.text.y  = element_text(size = 9),
    plot.title   = element_text(hjust = 0.5, face = "bold")
  ) +
  labs(title = "Spearman Correlation – Cscore (all samples)", x = "", y = "")
print(spear_p)
ggsave(file.path(plots_dir, "spearman_corr.pdf"), spear_p, width = 10, height = 6)
```

## 3.3 Barplots

```{r QC barplots}
# ============================================================================
# QC barplot — total number of bins per sample
# ============================================================================

sample_comp <- bind_rows(
  data.frame(Sample    = "REH",
             Condition = "REH",
             Total     = nrow(corrected$REH)),
  data.frame(Sample    = "EP1_1",
             Condition = "EP1",
             Total     = nrow(corrected$EP1_1)),
  data.frame(Sample    = "EP1_2",
             Condition = "EP1",
             Total     = nrow(corrected$EP1_2)),
  data.frame(Sample    = "EP1_merge",
             Condition = "EP1",
             Total     = nrow(corrected$EP1_merge)),
  data.frame(Sample    = "EP1_Aro_1",
             Condition = "EP1_Aro",
             Total     = nrow(corrected$EP1_Aro_1)),
  data.frame(Sample    = "EP1_Aro_2",
             Condition = "EP1_Aro",
             Total     = nrow(corrected$EP1_Aro_2)),
  data.frame(Sample    = "EP1_Aro_merge",
             Condition = "EP1_Aro",
             Total     = nrow(corrected$EP1_Aro_merge))
) %>%
  mutate(Sample = factor(Sample, levels = c("REH",
                                             "EP1_1", "EP1_2", "EP1_merge",
                                             "EP1_Aro_1", "EP1_Aro_2", "EP1_Aro_merge")))

bins_QC <- ggplot(sample_comp, aes(x = Sample, y = Total, fill = Condition)) +
  geom_bar(stat      = "identity",
           color     = "black",
           linewidth = 0.35,
           width     = 0.75) +
  geom_text(aes(label = scales::comma(Total)),
            vjust    = -0.4,
            size     = 3.5,
            fontface = "bold") +
  scale_fill_manual(values = cond_colors, name = "Condition") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)),
                     labels = scales::comma) +
  theme_bw() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank(),
    axis.text.x        = element_text(size = 11, angle = 35,
                                      hjust = 1, face = "bold"),
    axis.text.y        = element_text(size = 10),
    axis.title         = element_text(size = 13),
    plot.title         = element_text(hjust = 0.5, face = "bold", size = 14),
    legend.title       = element_text(face = "bold")
  ) +
  xlab("") + ylab("Number of bins") +
  labs(title = "Total bins per sample (QC)")
print(bins_QC)
ggsave(file.path(plots_dir, "bins_x_sample_QC.pdf"), bins_QC, width = 10, height = 6)
```
---

# 4. A/B Compartment classification

> **All downstream analyses use the corrected merged data (`cs_REH_corr`).**

Bins with **cscore > 0** are assigned to **compartment A** (open, active chromatin)
and bins with **cscore ≤ 0** to **compartment B** (closed, inactive chromatin).
Classification is performed on the **merged** samples only for downstream
biological interpretation.

```{r compartment-classification}
cs_REH_merge <- cs_REH_corr %>%
  select(chr, start, end,
         score_REH, score_EP1_merge, score_EP1_Aro_merge) %>%
  mutate(
    comp_REH     = ifelse(score_REH > 0,             "A", "B"),
    comp_EP1     = ifelse(score_EP1_merge > 0,       "A", "B"),
    comp_EP1_Aro = ifelse(score_EP1_Aro_merge > 0,   "A", "B")
  )

cs_REH_merge <- cs_REH_merge %>%
  mutate(
    dyn_REH_EP1    = paste(comp_REH, comp_EP1,     sep = "_"),
    dyn_EP1_EP1Aro = paste(comp_EP1, comp_EP1_Aro, sep = "_"),
    dyn_all        = paste(comp_REH, comp_EP1, comp_EP1_Aro, sep = "_")
  )

# Summary table
summary_comp <- data.frame(
  Condition = c("REH", "EP1", "EP1_Aro"),
  A_bins    = c(sum(cs_REH_merge$comp_REH == "A"),
                sum(cs_REH_merge$comp_EP1 == "A"),
                sum(cs_REH_merge$comp_EP1_Aro == "A")),
  B_bins    = c(sum(cs_REH_merge$comp_REH == "B"),
                sum(cs_REH_merge$comp_EP1 == "B"),
                sum(cs_REH_merge$comp_EP1_Aro == "B"))
) %>%
  mutate(
    Total   = A_bins + B_bins,
    Pct_A   = round(A_bins / Total * 100, 1),
    Pct_B   = round(B_bins / Total * 100, 1)
  )

knitr::kable(summary_comp, caption = "A/B compartment counts per condition (merged samples)")
write.table(summary_comp,
            file.path(files_dir, "summary_comp.tsv"),
            sep = "\t", row.names = FALSE, quote = FALSE)
```

## 4.1 A/B barplot

```{r ab-barplot}
dt_merge <- rbind(
  table(cs_REH_merge$comp_REH),
  table(cs_REH_merge$comp_EP1),
  table(cs_REH_merge$comp_EP1_Aro)
)
rownames(dt_merge) <- c("REH", "EP1", "EP1_Aro")

AB_barp <- ggplot(melt(dt_merge), aes(fill = Var1, x = Var2, y = value)) +
  geom_bar(position = "dodge", stat = "identity",
           width = 0.7, color = "black") +
  scale_fill_manual(values = c("REH" = "#e74c3c",
                               "EP1" = "#3498db",
                               "EP1_Aro" = "#2ecc71")) +
  theme_bw() +
  theme(
    panel.grid   = element_blank(),
    axis.text    = element_text(size = 12, face = "bold"),
    axis.title   = element_text(size = 14),
    plot.title   = element_text(size = 15, face = "bold", hjust = 0.5),
    legend.title = element_blank()
  ) +
  xlab("") + ylab("Number of bins") +
  labs(title = "A/B compartments – merged samples")
print(AB_barp)
ggsave(file.path(plots_dir, "AB_counts.pdf"), AB_barp, width = 10, height = 6)
```

```{r Proportion barplot}
# Build proportion data from all samples (replicates + merges)
prop_df <- data.frame(
  Sample = c("REH", "EP1_1", "EP1_2", "EP1_Aro_1", "EP1_Aro_2", "EP1_merge", "EP1_Aro_merge"),
  A = c(sum(cs_REH_merge$comp_REH == "A"),
        sum(corrected$EP1_1$score_EP1_1 > 0),
        sum(corrected$EP1_2$score_EP1_2 > 0),
        sum(corrected$EP1_Aro_1$score_EP1_Aro_1 > 0),
        sum(corrected$EP1_Aro_2$score_EP1_Aro_2 > 0),
        sum(cs_REH_merge$comp_EP1 == "A"),
        sum(cs_REH_merge$comp_EP1_Aro == "A")),
  B = c(sum(cs_REH_merge$comp_REH == "B"),
        sum(corrected$EP1_1$score_EP1_1 <= 0),
        sum(corrected$EP1_2$score_EP1_2 <= 0),
        sum(corrected$EP1_Aro_1$score_EP1_Aro_1 <= 0),
        sum(corrected$EP1_Aro_2$score_EP1_Aro_2 <= 0),
        sum(cs_REH_merge$comp_EP1 == "B"),
        sum(cs_REH_merge$comp_EP1_Aro == "B"))
) %>%
  mutate(Total  = A + B,
         Pct_A  = A / Total * 100,
         Pct_B  = B / Total * 100) %>%
  pivot_longer(cols = c(Pct_A, Pct_B),
               names_to  = "Compartment",
               values_to = "Proportion") %>%
  mutate(Compartment = recode(Compartment, "Pct_A" = "A", "Pct_B" = "B"),
         Sample = factor(Sample, levels = c("REH","EP1_1","EP1_2",
                                             "EP1_merge","EP1_Aro_1",
                                             "EP1_Aro_2","EP1_Aro_merge")))

prop_bar <- ggplot(prop_df, aes(x = Sample, y = Proportion, fill = Compartment)) +
  geom_bar(stat = "identity", color = "black", linewidth = 0.4) +
  geom_text(aes(label = paste0(round(Proportion, 1), "%")),
            position = position_stack(vjust = 0.5),
            size = 3.8, fontface = "bold", color = "white") +
  scale_fill_manual(values = c("A" = "#f0a500", "B" = "#6c3483")) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 101)) +
  theme_bw() +
  theme(
    panel.grid   = element_blank(),
    axis.text.x  = element_text(size = 11, angle = 30, hjust = 1, face = "bold"),
    axis.text.y  = element_text(size = 10),
    axis.title   = element_text(size = 13),
    plot.title   = element_text(size = 14, face = "bold", hjust = 0.5),
    legend.title = element_text(face = "bold")
  ) +
  xlab("") + ylab("Proportion (%)") +
  labs(title = "A/B compartment proportions per sample")
print(prop_bar)
ggsave(file.path(plots_dir, "AB_proportions.pdf"), prop_bar, width = 10, height = 6)
```

---

# 5. Compartment switching patterns across experimental conditions

Although the conditions are shown in the order REH → EP1 → EP1_Aro for visualization purposes, these samples represent independent transfections rather than a true temporal series. Therefore, the switching classes describe compartment state combinations across conditions, not a biological time course.

Each genomic bin is tracked across all three conditions and assigned one of
**8 switching classes** (2³ combinations of A/B states, named as REH_EP1_EP1Aro).

| Switching | Interpretation |
|-----------|----------------|
| A_A_A | Constitutively active |
| B_B_B | Constitutively inactive |
| B_A_A | Activation at EP1 variants |
| B_B_A | Aro-specific activation |
| A_B_B | Inactivation at EP1 variants |
| A_A_B | Aro-specific inactivation |
| A_B_A | EP1-specific inactivation |
| B_A_B | EP1-specific activation |

## 5.1 Switching classification

```{r switching-classification}
# Define switching levels and colors keyed directly on A/B state codes
switching_levels <- c(
  "A_A_A", "B_B_B",   # constitutive
  "B_A_A", "B_B_A",   # gained A
  "A_B_B", "A_A_B",   # gained B
  "A_B_A", "B_A_B"    # transient
)

switching_colors <- c(
  "A_A_A" = "#2ecc71",   # green        – constitutive A
  "B_B_B" = "#3498db",   # blue         – constitutive B
  "B_A_A" = "#e67e22",   # orange       – B→A at EP1, maintained
  "B_B_A" = "#f1c40f",   # yellow       – B→A only at EP1_Aro
  "A_B_B" = "#8e44ad",   # purple       – A→B at EP1, maintained
  "A_A_B" = "#e74c3c",   # red          – A→B only at EP1_Aro
  "A_B_A" = "#c0392b",   # dark red     – A→B→A (EP1-specific B)
  "B_A_B" = "#1abc9c"    # teal         – B→A→B (EP1-specific A)
)

cs_REH_merge$switching <- factor(cs_REH_merge$dyn_all,
                                   levels = switching_levels)

switch_counts <- as.data.frame(table(switching = cs_REH_merge$switching)) %>%
  mutate(
    Pct   = round(Freq / nrow(cs_REH_merge) * 100, 2),
    Group = case_when(
      switching %in% c("A_A_A", "B_B_B") ~ "Constitutive",
      switching %in% c("B_A_A", "B_B_A") ~ "Gain A",
      switching %in% c("A_B_B", "A_A_B") ~ "Gain B",
      TRUE                                 ~ "Transient"
    )
  )

knitr::kable(switch_counts[, c("switching", "Freq", "Pct", "Group")],
             col.names = c("Switching (REH_EP1_EP1Aro)", "Bins", "% total", "Group"),
             caption   = "Bin counts per switching class across REH → EP1 → EP1_Aro")
write.table(switch_counts,
            file.path(files_dir, "switch_counts.tsv"),
            sep = "\t", row.names = FALSE, quote = FALSE)
```

## 5.2 Bar plot of switching counts

```{r switching-barplot, fig.width=10, fig.height=5}
# Exclude constitutive bins for a clearer view of dynamics
switch_dynamic <- switch_counts[!switch_counts$switching %in% c("A_A_A", "B_B_B"), ]

switch_bar <- ggplot(switch_dynamic,
       aes(x    = reorder(switching, -Freq),
           y    = Freq,
           fill = switching)) +
  geom_bar(stat = "identity", color = "black", width = 0.7) +
  geom_text(aes(label = paste0(Pct, "%")),
            vjust = -0.4, size = 3.5, fontface = "bold") +
  scale_fill_manual(values = switching_colors) +
  theme_bw() +
  theme(
    panel.grid      = element_blank(),
    axis.text.x     = element_text(size = 11, angle = 30, hjust = 1, face = "bold"),
    axis.text.y     = element_text(size = 11),
    axis.title      = element_text(size = 13),
    plot.title      = element_text(size = 15, face = "bold", hjust = 0.5),
    legend.position = "none"
  ) +
  xlab("Switching class (REH_EP1_EP1Aro)") + ylab("Number of bins") +
  labs(title = "Compartment switching patterns across conditions (REH, EP1, EP1_Aro)\n(constitutive bins excluded)")
print(switch_bar)
ggsave(file.path(plots_dir, "switching_barplot.pdf"), switch_bar, width = 10, height = 6)
```

## 5.3 Alluvial / Sankey plot of transitions

The alluvial plot traces each bin's compartment state at each condition,
making it easy to see the dominant flows between states.

```{r Alluvial / Sankey plot}
library(ggalluvial)

alluv_sum <- cs_REH_merge %>%
  select(switching,
         REH     = comp_REH,
         EP1     = comp_EP1,
         EP1_Aro = comp_EP1_Aro) %>%
  pivot_longer(cols      = c(REH, EP1, EP1_Aro),
               names_to  = "Condition",
               values_to = "Compartment") %>%
  mutate(
    Condition = factor(Condition, levels = c("REH", "EP1", "EP1_Aro")),
    switching = factor(switching, levels = switching_levels)
  ) %>%
  group_by(switching, Condition, Compartment) %>%
  summarise(Freq = n(), .groups = "drop")

alluv_p <- ggplot(alluv_sum,
       aes(x        = Condition,
           y        = Freq,
           stratum  = Compartment,
           alluvium = switching,
           fill     = switching,
           label    = Compartment)) +
  geom_alluvium(alpha     = 0.55,
                color     = NA,
                linewidth = 0) +
  geom_stratum(alpha     = 0.88,
               color     = "white",
               linewidth = 0.6) +
  geom_text(stat     = "stratum",
            size     = 5,
            fontface = "bold",
            color    = "white") +
  scale_fill_manual(values = switching_colors,
                    name   = "Switching\n(REH_EP1_EP1Aro)") +
  scale_y_continuous(expand = c(0, 0)) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid      = element_blank(),
    axis.text.x     = element_text(size = 14, face = "bold", color = "gray20"),
    axis.text.y     = element_text(size = 10, color = "gray40"),
    axis.title.y    = element_text(size = 13, color = "gray30"),
    plot.title      = element_text(size = 15, face = "bold", hjust = 0.5,
                                   margin = margin(b = 12)),
    plot.background = element_rect(fill = "gray97", color = NA),
    legend.text     = element_text(size = 9),
    legend.title    = element_text(size = 10, face = "bold")
  ) +
  xlab("") + ylab("Number of bins") +
  labs(title = "Compartment state distribution across conditions (REH, EP1, EP1_Aro)")
print(alluv_p)
ggsave(file.path(plots_dir, "switch_alluvian.pdf"), alluv_p, width = 10, height = 6)
```

## 5.4 Pie chart of dynamic bins only

Among bins that change state at least once across the three conditions.

```{r pie-dynamic, fig.width=7, fig.height=7}
dynamic_bins <- cs_REH_merge %>%
  filter(!switching %in% c("A_A_A", "B_B_B"))

n_dynamic <- nrow(dynamic_bins)
n_total   <- nrow(cs_REH_merge)

pie_df <- as.data.frame(table(switching = dynamic_bins$switching)) %>%
  filter(Freq > 0) %>%
  mutate(
    Pct      = round(Freq / n_dynamic * 100, 1),
    PieLabel = paste0(Pct, "%")
  )

switch_pie <- ggplot(pie_df, aes(x = "", y = Freq, fill = switching)) +
  geom_col(width = 1, color = "black", linewidth = 1.1) +
  geom_text(aes(label = paste0(switching, "\n", PieLabel)),
            position = position_stack(vjust = 0.5),
            size = 3.5, fontface = "bold", color = "white") +
  coord_polar(theta = "y", start = 0) +
  scale_fill_manual(values = switching_colors,
                    name   = "Switching\n(REH_EP1_EP1Aro)") +
  theme_void() +
  ggtitle(paste0("Dynamic bins: n = ", n_dynamic,
                 " (", round(n_dynamic / n_total * 100, 1), "% of all bins)")) +
  theme(
    plot.title      = element_text(hjust = 0.5, size = 13, face = "bold"),
    legend.text     = element_text(size = 10),
    legend.title    = element_text(size = 11, face = "bold"),
    legend.position = "right"
  )
print(switch_pie)
ggsave(file.path(plots_dir, "switching_pie.pdf"), switch_pie, width = 10, height = 6)
```

## 5.5 Summary: switching groups

```{r switching-summary-table}
group_summary <- switch_counts %>%
  group_by(Group) %>%
  summarise(
    Bins             = sum(Freq),
    Pct_total        = round(sum(Freq) / nrow(cs_REH_merge) * 100, 1),
    Switching_classes = paste(switching, collapse = ", ")
  ) %>%
  arrange(desc(Bins))

knitr::kable(group_summary,
             col.names = c("Group", "Bins", "% of total", "Switching classes"),
             caption   = "Compartment switching groups – REH → EP1 → EP1_Aro")
write.table(group_summary,
            file.path(files_dir, "switch_groups.tsv"),
            sep = "\t", row.names = FALSE, quote = FALSE)
```

## 5.6 Heatmap

```{r heatmap, fig.width=12, fig.height=16, fig.align="center"}
# Filter dynamic bins and complete cases
dynamic_bins_complete <- cs_REH_merge %>%
  filter(!switching %in% c("A_A_A", "B_B_B")) %>%
  filter(!is.na(score_REH) & !is.na(score_EP1_merge) & !is.na(score_EP1_Aro_merge)) %>%
  arrange(switching)

# Prepare heatmap matrix
cscore_heat_dyn <- as.matrix(dynamic_bins_complete[, c("score_REH", "score_EP1_merge", "score_EP1_Aro_merge")])
rownames(cscore_heat_dyn) <- rownames(dynamic_bins_complete)
colnames(cscore_heat_dyn) <- c("REH", "EP1", "EP1_Aro")

# Prepare row annotation using switching codes directly
annot_row <- data.frame(
  Switching = factor(dynamic_bins_complete$switching,
                     levels = switching_levels[!switching_levels %in% c("A_A_A", "B_B_B")])
)
rownames(annot_row) <- rownames(cscore_heat_dyn)

# Annotation colors keyed on switching codes
annot_colors <- list(
  Switching = switching_colors[levels(annot_row$Switching)]
)

# Heatmap colors
heat_colors <- colorRampPalette(c("#2166AC", "white", "#B2182B"))(1000)

# Plot heatmap
pdf(file.path(plots_dir, "cscore_heatmap.pdf"), width = 10, height = 14)
pheatmap(cscore_heat_dyn,
         color                = heat_colors,
         cluster_rows         = FALSE,
         cluster_cols         = FALSE,
         show_rownames        = FALSE,
         annotation_row       = annot_row,
         annotation_colors    = annot_colors,
         annotation_names_row = FALSE,
         treeheight_row       = 50,
         scale                = "row",
         fontsize             = 12,
         fontsize_col         = 13,
         angle_col            = 45,
         legend_breaks        = c(-1, -0.5, 0, 0.5, 1),
         main                 = "Cscore Heatmap – Dynamic bins only (REH / EP1 / EP1_Aro)",
         mar                  = c(4, 6, 4, 10))
dev.off()

pheatmap(cscore_heat_dyn,
         color                = heat_colors,
         cluster_rows         = FALSE,
         cluster_cols         = FALSE,
         show_rownames        = FALSE,
         annotation_row       = annot_row,
         annotation_colors    = annot_colors,
         annotation_names_row = FALSE,
         treeheight_row       = 50,
         scale                = "row",
         fontsize             = 12,
         fontsize_col         = 13,
         angle_col            = 45,
         legend_breaks        = c(-1, -0.5, 0, 0.5, 1),
         main                 = "Cscore Heatmap – Dynamic bins only (REH / EP1 / EP1_Aro)",
         mar                  = c(4, 6, 4, 10))

```

---

# 6. Gene enrichment analysis on dynamic compartments

```{r Enrichment analysis}
# ============================================================================
# Compartment Switching – Enrichment Analysis (GenomicRanges + enrichR)
# ============================================================================
library(GenomicRanges)
library(enrichR)
library(dplyr)
library(ggplot2)

# ----------------------------------------------------------------------------
# 1. Load gene annotation BED (gencode v28)
# ----------------------------------------------------------------------------
genes_bed <- read.table("R:/R_STIK/xavi_scripts/genes_gencode.v28.bed",
                        sep       = "\t",
                        header    = FALSE,
                        col.names = c("chrom", "start", "end", "strand",
                                      "gene_name", "ID", "description")) %>%
  select(chrom, start, end, gene_name) %>%
  mutate(start = as.integer(start),
         end   = as.integer(end))

genes_gr <- makeGRangesFromDataFrame(genes_bed,
                                     seqnames.field     = "chrom",
                                     start.field        = "start",
                                     end.field          = "end",
                                     keep.extra.columns = TRUE)

# ----------------------------------------------------------------------------
# 2. Helper: get gene symbols overlapping a set of bins
# ----------------------------------------------------------------------------
get_genes_gr <- function(sw_class) {

  bins <- cs_REH_merge %>%
    filter(switching == sw_class) %>%
    select(chrom = chr, start, end) %>%
    mutate(start = as.integer(start),
           end   = as.integer(end))

  if (nrow(bins) == 0) return(character(0))

  bins_gr <- makeGRangesFromDataFrame(bins,
                                      seqnames.field = "chrom",
                                      start.field    = "start",
                                      end.field      = "end")

  hits <- findOverlaps(bins_gr, genes_gr)
  unique(genes_gr$gene_name[subjectHits(hits)])
}

# ----------------------------------------------------------------------------
# 3. Build gene list for all 6 dynamic switching classes
# ----------------------------------------------------------------------------
switching_dynamic_levels <- c("B_A_A", "B_B_A", "A_B_B", "A_A_B", "A_B_A", "B_A_B")

class_counts <- table(cs_REH_merge$switching)
cat("\nBin counts per dynamic switching class:\n")
print(class_counts[switching_dynamic_levels])

missing <- switching_dynamic_levels[
  !switching_dynamic_levels %in% names(class_counts) |
   class_counts[switching_dynamic_levels] == 0
]
if (length(missing) > 0) {
  cat("\nWARNING: These classes have 0 bins and will be skipped:", paste(missing, collapse = ", "), "\n")
} else {
  cat("\nAll 6 dynamic classes present in data. Proceeding with gene overlap.\n")
}

switching_gene_lists <- lapply(
  setNames(switching_dynamic_levels, switching_dynamic_levels),
  get_genes_gr
)

cat("\nGenes overlapping each switching class:\n")
gene_summary <- data.frame(
  switching = switching_dynamic_levels,
  n_genes   = sapply(switching_gene_lists, length)
)
print(gene_summary)

too_few <- gene_summary[gene_summary$n_genes < 5, ]
if (nrow(too_few) > 0) {
  cat("\nWARNING: The following classes have <5 genes and will be skipped by enrichR:\n")
  print(too_few)
}

# ----------------------------------------------------------------------------
# 4. Databases
# ----------------------------------------------------------------------------
dbs <- c("GO_Biological_Process_2023",
         "KEGG_2021_Human",
         "WikiPathway_2023_Human",
         "Reactome_2022",
         "MSigDB_Hallmark_2020")

websiteLive <- getOption("enrichR.live", TRUE)
setEnrichrSite("Enrichr")

# ----------------------------------------------------------------------------
# 5. Helpers
# ----------------------------------------------------------------------------
process_enrichr_db <- function(db_result) {
  df <- db_result %>%
    separate(col = "Overlap", into = c("Gene_count", "total"), sep = "/",
             convert = TRUE, extra = "drop", fill = "right")
  df$gene_ratio <- df$Gene_count / df$total
  df$Term <- gsub("\\s*\\(GO:\\d+\\)", "",  df$Term)
  df$Term <- gsub("\\s*R-HSA-[0-9.]+",  "", df$Term)
  df$Term <- gsub("\\s*WP\\d+",         "", df$Term)
  df$Term <- trimws(df$Term)
  df
}

save_enrichr_genelists <- function(sig_terms, prefix, out_dir) {
  sig_out <- sig_terms %>%
    select(Term, Gene_count, total, gene_ratio, P.value, Adjusted.P.value, Genes) %>%
    arrange(Adjusted.P.value)
  write.table(sig_out,
              file.path(out_dir, paste0(prefix, "_enrichment.tsv")),
              sep = "\t", row.names = FALSE, quote = FALSE)
}

plot_enrichr <- function(sig_terms, title_str, xlab_str) {
  plot_data <- head(sig_terms, 20)
  ggplot(plot_data, aes(x = Term, y = -log10(Adjusted.P.value))) +
    geom_hline(yintercept = 1, linetype = "dashed", color = "azure4", size = 0.5) +
    geom_point(aes(size = Gene_count, colour = gene_ratio), alpha = 0.7) +
    scale_x_discrete(limits = rev(plot_data$Term)) +
    scale_color_gradient(low = "green", high = "red", limits = c(0, NA)) +
    coord_flip() +
    theme_bw() +
    theme(axis.ticks.length  = unit(-0.1, "cm"),
          axis.text.x        = element_text(margin = margin(5, 5, 0, 5, "pt")),
          axis.text.y        = element_text(margin = margin(5, 5, 5, 5, "pt")),
          axis.text          = element_text(color = "black"),
          panel.grid.minor   = element_blank(),
          legend.title.align = 0.5,
          plot.title         = element_text(hjust = 0.5, color = "black",
                                            face = "bold", size = 14)) +
    xlab(xlab_str) +
    ylab("-log10(Adjusted P-value)") +
    labs(color = "Gene Ratio", size = "Number\nof genes") +
    ggtitle(title_str)
}

# ----------------------------------------------------------------------------
# 6. Main enrichR loop
# ----------------------------------------------------------------------------
enrichr_compartment_results <- list()

for (sw_class in switching_dynamic_levels) {

  gene_vec <- switching_gene_lists[[sw_class]]

  cat("\n=== Switching:", sw_class,
      "| Genes:", length(gene_vec), "===\n")

  if (length(gene_vec) < 5 || !websiteLive) {
    cat("  Skipping – too few genes or enrichR offline\n")
    next
  }

  enriched <- enrichr(gene_vec, dbs)
  enrichr_compartment_results[[sw_class]] <- enriched

  for (db_idx in seq_along(dbs)) {
    db_name <- dbs[db_idx]

    if (is.null(enriched[[db_idx]]) || nrow(enriched[[db_idx]]) == 0) next

    enriched_processed <- process_enrichr_db(enriched[[db_idx]])
    sig_terms          <- subset(enriched_processed, Adjusted.P.value < 0.05)

    if (nrow(sig_terms) == 0) {
      cat("  No significant terms:", db_name, "\n")
      next
    }

    save_enrichr_genelists(
      sig_terms,
      prefix  = paste0(sw_class, "_", db_name),
      out_dir = files_dir
    )

    p <- plot_enrichr(
      sig_terms,
      title_str = paste0(db_name, "\nSwitching: ", sw_class),
      xlab_str  = paste0(db_name, " Terms")
    )
    print(p)

    ggsave(
      filename = file.path(plots_dir,
                           paste0("enrichR_", db_name, "_", sw_class, ".pdf")),
      plot   = p,
      width  = 10,
      height = 7
    )

    cat("  Saved:", db_name, "\n")
  }
}

cat("\n=== Compartment enrichR analysis complete ===\n")
```

---

# 7. ChIP-Seq EP1 intersection

```{r ChIP-Seq intersection}
# ============================================================================
# Intersection of switching compartments with ChIP-seq peaks
# ============================================================================
library(GenomicRanges)
library(ggplot2)
library(dplyr)
library(tidyr)

# ----------------------------------------------------------------------------
# Load ChIP-seq peaks BED
# ----------------------------------------------------------------------------
peaks_bed <- read.table("R:/R_STIK/chip_seq/Xavi_TFM/697HF_E2APBX1-HF_peaks.bed",
                        sep       = "\t",
                        header    = FALSE,
                        col.names = c("chrom", "start", "end")
                        [1:ncol(read.table(
                        "R:/R_STIK/chip_seq/Xavi_TFM/697HF_E2APBX1-HF_peaks.bed",
                        sep = "\t", nrows = 1))]) %>%
  select(chrom = 1, start = 2, end = 3) %>%
  mutate(start = as.integer(start),
         end   = as.integer(end))

###Put "R:/R_STIK/" or "/ijc/LABS/STIK/RAW" depending on the root structure

peaks_gr <- makeGRangesFromDataFrame(peaks_bed,
                                     seqnames.field = "chrom",
                                     start.field    = "start",
                                     end.field      = "end")

cat("Total peaks loaded:", length(peaks_gr), "\n")

# ----------------------------------------------------------------------------
# Overlap peaks with every switching class
# ----------------------------------------------------------------------------
all_switching_levels <- c("A_A_A", "B_B_B",
                           "B_A_A", "B_B_A",
                           "A_B_B", "A_A_B",
                           "A_B_A", "B_A_B")

overlap_counts <- lapply(all_switching_levels, function(sw) {

  bins <- cs_REH_merge %>%
    filter(switching == sw) %>%
    select(chrom = chr, start, end) %>%
    mutate(start = as.integer(start),
           end   = as.integer(end))

  bins_gr <- makeGRangesFromDataFrame(bins,
                                      seqnames.field = "chrom",
                                      start.field    = "start",
                                      end.field      = "end")

  n_bins      <- length(bins_gr)
  n_overlap   <- length(unique(queryHits(findOverlaps(bins_gr, peaks_gr))))
  pct_overlap <- n_overlap / n_bins * 100

  data.frame(switching   = sw,
             n_bins      = n_bins,
             n_overlap   = n_overlap,
             pct_overlap = pct_overlap)
}) %>% bind_rows()

# ----------------------------------------------------------------------------
# Background: constitutive bins only (A_A_A + B_B_B)
# ----------------------------------------------------------------------------
constitutive_counts <- overlap_counts %>%
  # filter(switching %in% c("A_A_A", "B_B_B")) %>%
  summarise(n_bins    = sum(n_bins),
            n_overlap = sum(n_overlap))

bg_bins_with    <- constitutive_counts$n_overlap
bg_bins_without <- constitutive_counts$n_bins - constitutive_counts$n_overlap
background_pct  <- bg_bins_with / constitutive_counts$n_bins * 100

cat(sprintf("Constitutive background: %d bins, %d with peak (%.2f%%)\n",
            constitutive_counts$n_bins, bg_bins_with, background_pct))

# ----------------------------------------------------------------------------
# Fisher's exact test for each switching class vs constitutive background
# ----------------------------------------------------------------------------
# switching_classes <- setdiff(all_switching_levels, c("A_A_A", "B_B_B"))
switching_classes <- all_switching_levels

fisher_results <- lapply(switching_classes, function(sw) {

  sw_row <- overlap_counts %>% filter(switching == sw)

  sw_with    <- sw_row$n_overlap
  sw_without <- sw_row$n_bins - sw_row$n_overlap

  # 2x2 contingency table:
  #              Peak    No Peak
  # Switching  [ sw_with,    sw_without ]
  # Background [ bg_with,    bg_without ]
  ct <- matrix(c(sw_with,    sw_without,
                 bg_bins_with, bg_bins_without),
               nrow = 2, byrow = TRUE,
               dimnames = list(c("Switching", "Background"),
                               c("Peak", "No_Peak")))

ft <- fisher.test(ct, alternative = "two.sided")

  data.frame(
    switching  = sw,
    n_bins     = sw_row$n_bins,
    n_overlap  = sw_with,
    pct_overlap = sw_row$pct_overlap,
    OR         = ft$estimate,
    CI_low     = ft$conf.int[1],
    CI_high    = ft$conf.int[2],
    p_value    = ft$p.value
  )
}) %>% bind_rows()

# Multiple testing correction (BH)
fisher_results <- fisher_results %>%
  mutate(
    p_adj          = p.adjust(p_value, method = "BH"),
    significance   = case_when(
      p_adj < 0.001 ~ "***",
      p_adj < 0.01  ~ "**",
      p_adj < 0.05  ~ "*",
      TRUE          ~ "ns"
    ),
    fold_enrichment = pct_overlap / background_pct,
    switching       = factor(switching, levels = all_switching_levels)
  )

print(fisher_results[, c("switching", "n_bins", "n_overlap",
                          "pct_overlap", "fold_enrichment",
                          "OR", "CI_low", "CI_high",
                          "p_value", "p_adj", "significance")])

# Also add fold enrichment to full overlap_counts for plotting
overlap_counts <- overlap_counts %>%
  mutate(
    fold_enrichment = pct_overlap / background_pct,
    switching       = factor(switching, levels = all_switching_levels)
  )

# ----------------------------------------------------------------------------
# Annotate each bin with peak overlap for boxplot
# ----------------------------------------------------------------------------
all_bins_gr <- makeGRangesFromDataFrame(
  cs_REH_merge %>% select(chr, start, end),
  seqnames.field = "chr",
  start.field    = "start",
  end.field      = "end"
)

cs_REH_merge <- cs_REH_merge %>%
  mutate(
    peak_overlap = ifelse(countOverlaps(all_bins_gr, peaks_gr) > 0, "Peak", "No peak")
  )

# ----------------------------------------------------------------------------
# Plot 1 — Polar enrichment plot with all 8 switching classes
# (constitutive dashed reference line at background_pct-based fold = 1)
# ----------------------------------------------------------------------------

overlap_polar <- ggplot(overlap_counts,
                        aes(x = switching, y = fold_enrichment, fill = switching)) +
  geom_col(color = "black", linewidth = 0.4, width = 0.85) +
  geom_hline(yintercept = 1, linetype = "dashed",
             color = "gray40", linewidth = 0.5) +
  coord_polar() +
  scale_fill_manual(values = switching_colors,
                    name   = "Switching\n(REH_EP1_EP1Aro)") +
  scale_y_continuous(limits = c(0, max(overlap_counts$fold_enrichment) * 1.2)) +
  theme_bw() +
  theme(
    axis.text.x  = element_text(size = 8, face = "bold"),
    axis.text.y  = element_text(size = 8),
    axis.title.x = element_blank(),
    axis.title.y = element_text(size = 11),
    plot.title   = element_text(hjust = 0.5, face = "bold", size = 13),
    panel.grid   = element_line(color = "gray85", linetype = "dashed"),
    legend.title = element_text(face = "bold")
  ) +
  ylab("Fold Enrichment over constitutive background") +
  labs(title = "E2A-PBX1 peak enrichment per compartment switching class\n(REH_EP1_EP1Aro)")
print(overlap_polar)
ggsave(file.path(plots_dir, "EP1_overlap_polar.pdf"), overlap_polar, width = 10, height = 6)

# ----------------------------------------------------------------------------
# Plot 2 — Forest plot: Odds Ratio + 95% CI per switching class
# ----------------------------------------------------------------------------
forest_plot <- ggplot(fisher_results,
                      aes(x = OR, y = switching, color = switching)) +
  geom_vline(xintercept = 1, linetype = "dashed",
             color = "gray40", linewidth = 0.5) +
  geom_errorbarh(aes(xmin = CI_low, xmax = CI_high),
                 height = 0.25, linewidth = 0.7) +
  geom_point(size = 3.5) +
  geom_text(aes(label = significance, x = CI_high),
            hjust = -0.4, size = 5, color = "black") +
  scale_color_manual(values = switching_colors,
                     name   = "Switching class") +
  scale_x_log10() +
  theme_bw() +
  theme(
    axis.text.y  = element_text(size = 11, face = "bold"),
    axis.text.x  = element_text(size = 10),
    axis.title   = element_text(size = 12),
    plot.title   = element_text(hjust = 0.5, face = "bold", size = 13),
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank(),
    legend.position    = "none"
  ) +
  xlab("Odds Ratio (log scale) vs constitutive background") +
  ylab("Switching class") +
  labs(title = "Fisher's exact test: E2A-PBX1 peak enrichment\nvs constitutive bins (A_A_A + B_B_B)")
print(forest_plot)
ggsave(file.path(plots_dir, "EP1_fisher_forest.pdf"), forest_plot, width = 8, height = 5)

# ----------------------------------------------------------------------------
# Plot 3 — Boxplot: cscore per condition, faceted by peak overlap
# ----------------------------------------------------------------------------
box_df <- cs_REH_merge %>%
  select(peak_overlap,
         REH     = score_REH,
         EP1     = score_EP1_merge,
         EP1_Aro = score_EP1_Aro_merge) %>%
  pivot_longer(cols      = c(REH, EP1, EP1_Aro),
               names_to  = "Condition",
               values_to = "cscore") %>%
  mutate(
    Condition    = factor(Condition, levels = c("REH", "EP1", "EP1_Aro")),
    peak_overlap = factor(peak_overlap, levels = c("Peak", "No peak")),
    fill_group   = ifelse(peak_overlap == "Peak", as.character(Condition), "No peak")
  )

fill_colors <- c(cond_colors, "No peak" = "#aaaaaa")

cscore_EP1_box <- ggplot(box_df, aes(x = Condition, y = cscore, fill = fill_group)) +
  geom_boxplot(outlier.size  = 0.3,
               outlier.alpha = 0.3,
               linewidth     = 0.45,
               alpha         = 0.9) +
  geom_hline(yintercept = 0, linetype = "dashed",
             color = "gray30", linewidth = 0.4) +
  facet_wrap(~ peak_overlap, nrow = 1) +
  scale_fill_manual(values = fill_colors,
                    breaks = names(cond_colors),
                    name   = "Condition") +
  scale_y_continuous(limits = c(-1, 1),
                     breaks = c(-1, -0.5, 0, 0.5, 1)) +
  theme_bw() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank(),
    axis.text.x        = element_text(size = 11, angle = 30,
                                      hjust = 1, face = "bold"),
    axis.text.y        = element_text(size = 10),
    axis.title         = element_text(size = 13),
    strip.text         = element_text(size = 12, face = "bold"),
    strip.background   = element_rect(fill = "gray92", color = "black"),
    plot.title         = element_text(hjust = 0.5, face = "bold", size = 13),
    legend.title       = element_text(face = "bold", size = 10),
    legend.text        = element_text(size = 9)
  ) +
  xlab("Condition") + ylab("cscore") +
  labs(title = "cscore distribution – bins overlapping E2A-PBX1 peaks vs non-overlapping")
print(cscore_EP1_box)
ggsave(file.path(plots_dir, "EP1_overlap_cscore.pdf"), cscore_EP1_box, width = 10, height = 6)
```
```{r ChIP-Seq intersection Top25}
# ============================================================================
# Intersection of switching compartments with Top25 ChIP-seq peaks
# ============================================================================
library(GenomicRanges)
library(ggplot2)
library(dplyr)
library(tidyr)

# ----------------------------------------------------------------------------
# Load Top25 ChIP-seq peaks BED
# ----------------------------------------------------------------------------
peaks_top25_bed <- read.table(
  "R:/R_STIK/chip_seq/Xavi_TFM/top25_697HF_E2APBX1-HF_peaks.bed",
  sep    = "\t",
  header = FALSE
) %>%
  select(chrom = 1, start = 2, end = 3) %>%
  mutate(start = as.integer(start),
         end   = as.integer(end))

peaks_top25_gr <- makeGRangesFromDataFrame(peaks_top25_bed,
                                           seqnames.field = "chrom",
                                           start.field    = "start",
                                           end.field      = "end")

cat("Total top25 peaks loaded:", length(peaks_top25_gr), "\n")

# ----------------------------------------------------------------------------
# Overlap top25 peaks with every switching class
# ----------------------------------------------------------------------------
overlap_top25_counts <- lapply(all_switching_levels, function(sw) {

  bins <- cs_REH_merge %>%
    filter(switching == sw) %>%
    select(chrom = chr, start, end) %>%
    mutate(start = as.integer(start),
           end   = as.integer(end))

  bins_gr <- makeGRangesFromDataFrame(bins,
                                      seqnames.field = "chrom",
                                      start.field    = "start",
                                      end.field      = "end")

  n_bins      <- length(bins_gr)
  n_overlap   <- length(unique(queryHits(findOverlaps(bins_gr, peaks_top25_gr))))
  pct_overlap <- n_overlap / n_bins * 100

  data.frame(switching   = sw,
             n_bins      = n_bins,
             n_overlap   = n_overlap,
             pct_overlap = pct_overlap)
}) %>% bind_rows()

# ----------------------------------------------------------------------------
# Background: all bins (same as original)
# ----------------------------------------------------------------------------
constitutive_top25_counts <- overlap_top25_counts %>%
  summarise(n_bins    = sum(n_bins),
            n_overlap = sum(n_overlap))

bg_top25_bins_with    <- constitutive_top25_counts$n_overlap
bg_top25_bins_without <- constitutive_top25_counts$n_bins - constitutive_top25_counts$n_overlap
background_top25_pct  <- bg_top25_bins_with / constitutive_top25_counts$n_bins * 100

cat(sprintf("Top25 background: %d bins, %d with peak (%.2f%%)\n",
            constitutive_top25_counts$n_bins,
            bg_top25_bins_with,
            background_top25_pct))

# ----------------------------------------------------------------------------
# Fisher's exact test for each switching class vs background
# ----------------------------------------------------------------------------
fisher_top25_results <- lapply(all_switching_levels, function(sw) {

  sw_row     <- overlap_top25_counts %>% filter(switching == sw)
  sw_with    <- sw_row$n_overlap
  sw_without <- sw_row$n_bins - sw_row$n_overlap

  ct <- matrix(c(sw_with,              sw_without,
                 bg_top25_bins_with,   bg_top25_bins_without),
               nrow = 2, byrow = TRUE,
               dimnames = list(c("Switching", "Background"),
                               c("Peak", "No_Peak")))

  ft <- fisher.test(ct, alternative = "two.sided")

  data.frame(
    switching   = sw,
    n_bins      = sw_row$n_bins,
    n_overlap   = sw_with,
    pct_overlap = sw_row$pct_overlap,
    OR          = ft$estimate,
    CI_low      = ft$conf.int[1],
    CI_high     = ft$conf.int[2],
    p_value     = ft$p.value
  )
}) %>% bind_rows()

fisher_top25_results <- fisher_top25_results %>%
  mutate(
    p_adj           = p.adjust(p_value, method = "BH"),
    significance    = case_when(
      p_adj < 0.001 ~ "***",
      p_adj < 0.01  ~ "**",
      p_adj < 0.05  ~ "*",
      TRUE          ~ "ns"
    ),
    fold_enrichment = pct_overlap / background_top25_pct,
    switching       = factor(switching, levels = all_switching_levels)
  )

print(fisher_top25_results[, c("switching", "n_bins", "n_overlap",
                                "pct_overlap", "fold_enrichment",
                                "OR", "CI_low", "CI_high",
                                "p_value", "p_adj", "significance")])

overlap_top25_counts <- overlap_top25_counts %>%
  mutate(
    fold_enrichment = pct_overlap / background_top25_pct,
    switching       = factor(switching, levels = all_switching_levels)
  )

# ----------------------------------------------------------------------------
# Annotate each bin with top25 peak overlap for boxplot
# ----------------------------------------------------------------------------
cs_REH_merge <- cs_REH_merge %>%
  mutate(
    peak_top25_overlap = ifelse(
      countOverlaps(all_bins_gr, peaks_top25_gr) > 0, "Peak", "No peak"
    )
  )

# ----------------------------------------------------------------------------
# Plot 1 — Polar enrichment plot
# ----------------------------------------------------------------------------
overlap_top25_polar <- ggplot(overlap_top25_counts,
                              aes(x = switching, y = fold_enrichment,
                                  fill = switching)) +
  geom_col(color = "black", linewidth = 0.4, width = 0.85) +
  geom_hline(yintercept = 1, linetype = "dashed",
             color = "gray40", linewidth = 0.5) +
  coord_polar() +
  scale_fill_manual(values = switching_colors,
                    name   = "Switching\n(REH_EP1_EP1Aro)") +
  scale_y_continuous(
    limits = c(0, max(overlap_top25_counts$fold_enrichment) * 1.2)
  ) +
  theme_bw() +
  theme(
    axis.text.x  = element_text(size = 8, face = "bold"),
    axis.text.y  = element_text(size = 8),
    axis.title.x = element_blank(),
    axis.title.y = element_text(size = 11),
    plot.title   = element_text(hjust = 0.5, face = "bold", size = 13),
    panel.grid   = element_line(color = "gray85", linetype = "dashed"),
    legend.title = element_text(face = "bold")
  ) +
  ylab("Fold Enrichment over background") +
  labs(title = "Top25 E2A-PBX1 peak enrichment per compartment switching class\n(REH_EP1_EP1Aro)")
print(overlap_top25_polar)
ggsave(file.path(plots_dir, "EP1_top25_overlap_polar.pdf"),
       overlap_top25_polar, width = 10, height = 6)

# ----------------------------------------------------------------------------
# Plot 2 — Forest plot
# ----------------------------------------------------------------------------
forest_top25_plot <- ggplot(fisher_top25_results,
                            aes(x = OR, y = switching, color = switching)) +
  geom_vline(xintercept = 1, linetype = "dashed",
             color = "gray40", linewidth = 0.5) +
  geom_errorbarh(aes(xmin = CI_low, xmax = CI_high),
                 height = 0.25, linewidth = 0.7) +
  geom_point(size = 3.5) +
  geom_text(aes(label = significance, x = CI_high),
            hjust = -0.4, size = 5, color = "black") +
  scale_color_manual(values = switching_colors,
                     name   = "Switching class") +
  scale_x_log10() +
  theme_bw() +
  theme(
    axis.text.y        = element_text(size = 11, face = "bold"),
    axis.text.x        = element_text(size = 10),
    axis.title         = element_text(size = 12),
    plot.title         = element_text(hjust = 0.5, face = "bold", size = 13),
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank(),
    legend.position    = "none"
  ) +
  xlab("Odds Ratio (log scale) vs background") +
  ylab("Switching class") +
  labs(title = "Fisher's exact test: Top25 E2A-PBX1 peak enrichment\nvs all bins background")
print(forest_top25_plot)
ggsave(file.path(plots_dir, "EP1_top25_fisher_forest.pdf"),
       forest_top25_plot, width = 8, height = 5)

# ----------------------------------------------------------------------------
# Plot 3 — Boxplot: cscore per condition, faceted by top25 peak overlap
# ----------------------------------------------------------------------------
box_top25_df <- cs_REH_merge %>%
  select(peak_top25_overlap,
         REH     = score_REH,
         EP1     = score_EP1_merge,
         EP1_Aro = score_EP1_Aro_merge) %>%
  pivot_longer(cols      = c(REH, EP1, EP1_Aro),
               names_to  = "Condition",
               values_to = "cscore") %>%
  mutate(
    Condition          = factor(Condition, levels = c("REH", "EP1", "EP1_Aro")),
    peak_top25_overlap = factor(peak_top25_overlap, levels = c("Peak", "No peak")),
    fill_group         = ifelse(peak_top25_overlap == "Peak",
                                as.character(Condition), "No peak")
  )

cscore_top25_box <- ggplot(box_top25_df,
                           aes(x = Condition, y = cscore, fill = fill_group)) +
  geom_boxplot(outlier.size  = 0.3,
               outlier.alpha = 0.3,
               linewidth     = 0.45,
               alpha         = 0.9) +
  geom_hline(yintercept = 0, linetype = "dashed",
             color = "gray30", linewidth = 0.4) +
  facet_wrap(~ peak_top25_overlap, nrow = 1) +
  scale_fill_manual(values = fill_colors,
                    breaks = names(cond_colors),
                    name   = "Condition") +
  scale_y_continuous(limits = c(-1, 1),
                     breaks = c(-1, -0.5, 0, 0.5, 1)) +
  theme_bw() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank(),
    axis.text.x        = element_text(size = 11, angle = 30,
                                      hjust = 1, face = "bold"),
    axis.text.y        = element_text(size = 10),
    axis.title         = element_text(size = 13),
    strip.text         = element_text(size = 12, face = "bold"),
    strip.background   = element_rect(fill = "gray92", color = "black"),
    plot.title         = element_text(hjust = 0.5, face = "bold", size = 13),
    legend.title       = element_text(face = "bold", size = 10),
    legend.text        = element_text(size = 9)
  ) +
  xlab("Condition") + ylab("cscore") +
  labs(title = "cscore distribution – bins overlapping Top25 E2A-PBX1 peaks vs non-overlapping")
print(cscore_top25_box)
ggsave(file.path(plots_dir, "EP1_top25_overlap_cscore.pdf"),
       cscore_top25_box, width = 10, height = 6)
```

---

# 8. Save outputs

```{r save}
write.table(cs_REH_merge,
            file.path(files_dir, "cs_REH_merge.tab"),
            sep       = "\t",
            quote     = FALSE,
            col.names = TRUE,
            row.names = FALSE)

cat("Merged table saved to files/cs_REH_merge.tab\n")
```

---

# 9. Loop intersection with ChIP

```{r Loop + ChIP}
# ============================================================================
# Intersection of loop anchors with ChIP-seq peaks (full + Top25)
# ============================================================================
library(GenomicRanges)
library(ggplot2)
library(dplyr)
library(tidyr)

# ----------------------------------------------------------------------------
# Output directories – separate from compartment analysis
# ----------------------------------------------------------------------------
loops_base_dir  <- "R:/R_STIK/Hi-C/HiC_marta_Oct25/loops/intersected_loops/fisher"
loops_files_dir <- file.path(loops_base_dir, "files")
loops_plots_dir <- file.path(loops_base_dir, "plots")

for (d in c(loops_files_dir, loops_plots_dir)) {
  if (!dir.exists(d)) dir.create(d, recursive = TRUE)
}

# ----------------------------------------------------------------------------
# Genome file (same as bedtools)
# ----------------------------------------------------------------------------
genome_file <- "R:/R_STIK/xavi_scripts/hg38.chrom.sizes.txt"
# On server:
# genome_file <- "/mnt/beegfs/xvilchez/files/hg38.chrom.sizes.txt"

# ----------------------------------------------------------------------------
# 1. Load loop anchor BED files
# ----------------------------------------------------------------------------
anchors_dir <- "R:/R_STIK/Hi-C/HiC_marta_Oct25/loops/intersected_loops"

anchor_files <- list(
  "ARO_specific"    = "ARO_specific_only_05_10kb_exp30kb_anchors.bed",
  "EP1_specific"    = "EP1_specific_only_05_10kb_exp30kb_anchors.bed",
  "REH_specific"    = "REH_specific_only_05_10kb_exp30kb_anchors.bed",
  "shared_EP1_ARO"  = "shared_EP1_ARO_only_05_10kb_exp30kb_anchors.bed",
  "shared_ARO_REH"  = "shared_ARO_REH_only_05_10kb_exp30kb_anchors.bed",
  "shared_EP1_REH"  = "shared_EP1_REH_only_05_10kb_exp30kb_anchors.bed",
  "shared_all3"     = "shared_loops_all3_05_10kb_exp30kb_anchors.bed"
)

anchor_levels <- names(anchor_files)

# Color palette for anchor classes
anchor_colors <- c(
  "REH_specific"   = "#e74c3c",
  "EP1_specific"   = "#3498db",
  "ARO_specific"   = "#2ecc71",
  "shared_EP1_REH" = "#9b59b6",
  "shared_ARO_REH" = "#e67e22",
  "shared_EP1_ARO" = "#1abc9c",
  "shared_all3"    = "#34495e"
)

read_anchor <- function(filename) {
  df <- read.table(file.path(anchors_dir, filename),
                   sep    = "\t",
                   header = FALSE) %>%
    select(chrom = 1, start = 2, end = 3) %>%
    mutate(start = as.integer(start),
           end   = as.integer(end))
  makeGRangesFromDataFrame(df,
                           seqnames.field = "chrom",
                           start.field    = "start",
                           end.field      = "end")
}

anchors_gr_list <- lapply(anchor_files, read_anchor)

cat("Anchor sets loaded:\n")
for (nm in names(anchors_gr_list)) {
  cat(sprintf("  %-20s : %d anchors\n", nm, length(anchors_gr_list[[nm]])))
}

# ----------------------------------------------------------------------------
# Helper: bedtools-fisher-equivalent overlap stats (bp-level)
# ----------------------------------------------------------------------------
compute_overlap_stats <- function(anchors_gr_list, peaks_gr, genome_file) {

  # Read chromosome sizes
  chrom_sizes <- read.table(genome_file, sep = "\t", header = FALSE,
                             col.names = c("chrom", "size")) %>%
    filter(chrom %in% seqlevels(peaks_gr))

  genome_size <- sum(as.numeric(chrom_sizes$size))
  cat(sprintf("  Genome size: %.0f bp\n", genome_size))

  # Peaks: total bp covered (merged)
  peaks_reduced <- reduce(peaks_gr)
  peak_bp       <- sum(as.numeric(width(peaks_reduced)))
  non_peak_bp   <- genome_size - peak_bp
  cat(sprintf("  Peak bp: %.0f (%.4f%%)\n", peak_bp, peak_bp / genome_size * 100))

  # Per anchor class: bp overlapping peaks
  counts <- lapply(names(anchors_gr_list), function(nm) {
    gr            <- anchors_gr_list[[nm]]
    anchor_bp     <- sum(as.numeric(width(reduce(gr))))
    overlap_gr    <- intersect(reduce(gr), peaks_reduced)
    overlap_bp    <- sum(as.numeric(width(overlap_gr)))
    no_overlap_bp <- anchor_bp - overlap_bp
    pct_overlap   <- overlap_bp / anchor_bp * 100

    data.frame(anchor_class  = nm,
               n_anchors     = length(gr),
               anchor_bp     = anchor_bp,
               overlap_bp    = overlap_bp,
               no_overlap_bp = no_overlap_bp,
               pct_overlap   = pct_overlap)
  }) %>% bind_rows()

  # Fisher test: bp-level 2x2 vs genome background (scaled to avoid overflow)
  scale <- 1e6 / genome_size

  fisher_res <- lapply(names(anchors_gr_list), function(nm) {
    row  <- counts %>% filter(anchor_class == nm)
    a    <- round(row$overlap_bp)
    b    <- round(row$no_overlap_bp)
    c_bg <- round(peak_bp     * scale)
    d_bg <- round(non_peak_bp * scale)

    ct <- matrix(c(a,    b,
                   c_bg, d_bg),
                 nrow = 2, byrow = TRUE,
                 dimnames = list(c("Anchors", "Genome"),
                                 c("Peak_bp", "Non_peak_bp")))

    ft <- fisher.test(ct, alternative = "greater")

    data.frame(anchor_class  = nm,
               n_anchors     = row$n_anchors,
               anchor_bp     = row$anchor_bp,
               overlap_bp    = row$overlap_bp,
               pct_overlap   = row$pct_overlap,
               OR            = ft$estimate,
               CI_low        = ft$conf.int[1],
               CI_high       = ft$conf.int[2],
               p_value       = ft$p.value)
  }) %>% bind_rows()

  bg_pct <- peak_bp / genome_size * 100

  fisher_res <- fisher_res %>%
    mutate(
      p_adj           = p.adjust(p_value, method = "BH"),
      significance    = case_when(
        p_adj < 0.001 ~ "***",
        p_adj < 0.01  ~ "**",
        p_adj < 0.05  ~ "*",
        TRUE          ~ "ns"
      ),
      fold_enrichment = pct_overlap / bg_pct,
      anchor_class    = factor(anchor_class, levels = anchor_levels)
    )

  counts <- counts %>%
    mutate(fold_enrichment = pct_overlap / bg_pct,
           anchor_class    = factor(anchor_class, levels = anchor_levels))

  list(counts = counts, fisher = fisher_res, bg_pct = bg_pct)
}

# ----------------------------------------------------------------------------
# Helper: polar plot
# ----------------------------------------------------------------------------
plot_anchor_polar <- function(counts, title_str) {
  ggplot(counts, aes(x = anchor_class, y = fold_enrichment, fill = anchor_class)) +
    geom_col(color = "black", linewidth = 0.4, width = 0.85) +
    geom_hline(yintercept = 1, linetype = "dashed",
               color = "gray40", linewidth = 0.5) +
    coord_polar() +
    scale_fill_manual(values = anchor_colors, name = "Anchor class") +
    scale_y_continuous(limits = c(0, max(counts$fold_enrichment) * 1.2)) +
    theme_bw() +
    theme(
      axis.text.x  = element_text(size = 8, face = "bold"),
      axis.text.y  = element_text(size = 8),
      axis.title.x = element_blank(),
      axis.title.y = element_text(size = 11),
      plot.title   = element_text(hjust = 0.5, face = "bold", size = 13),
      panel.grid   = element_line(color = "gray85", linetype = "dashed"),
      legend.title = element_text(face = "bold")
    ) +
    ylab("Fold Enrichment over genome background") +
    labs(title = title_str)
}

# ----------------------------------------------------------------------------
# Helper: forest plot
# ----------------------------------------------------------------------------
plot_anchor_forest <- function(fisher_res, title_str) {
  ggplot(fisher_res, aes(x = OR, y = anchor_class, color = anchor_class)) +
    geom_vline(xintercept = 1, linetype = "dashed",
               color = "gray40", linewidth = 0.5) +
    geom_errorbarh(aes(xmin = CI_low,
                       xmax = ifelse(is.infinite(CI_high), OR * 10, CI_high)),
                   height = 0.25, linewidth = 0.7) +
    geom_point(size = 3.5) +
    geom_text(aes(label = significance),
              hjust = -0.8, vjust = 0.4, size = 5, color = "black") +
    scale_color_manual(values = anchor_colors, name = "Anchor class") +
    scale_x_log10() +
    theme_bw() +
    theme(
      axis.text.y        = element_text(size = 11, face = "bold"),
      axis.text.x        = element_text(size = 10),
      axis.title         = element_text(size = 12),
      plot.title         = element_text(hjust = 0.5, face = "bold", size = 13),
      panel.grid.major.y = element_blank(),
      panel.grid.minor   = element_blank(),
      legend.position    = "none"
    ) +
    xlab("Odds Ratio (log scale) vs genome background") +
    ylab("Anchor class") +
    labs(title = title_str)
}

# ============================================================================
# A. Full 697HF E2A-PBX1 peaks
# ============================================================================
cat("\n--- Full E2A-PBX1 peaks ---\n")
res_full <- compute_overlap_stats(anchors_gr_list, peaks_gr, genome_file)

print(res_full$fisher[, c("anchor_class", "n_anchors", "overlap_bp",
                           "pct_overlap", "fold_enrichment",
                           "OR", "CI_low", "CI_high", "p_value", "p_adj", "significance")])

write.table(res_full$fisher,
            file.path(loops_files_dir, "loop_anchor_EP1_chip_fisher.tsv"),
            sep = "\t", row.names = FALSE, quote = FALSE)

p_polar_full <- plot_anchor_polar(
  res_full$counts,
  "E2A-PBX1 peak enrichment per loop anchor class\n(genome background)"
)
print(p_polar_full)
ggsave(file.path(loops_plots_dir, "loop_anchor_EP1_polar.pdf"),
       p_polar_full, width = 10, height = 6)

p_forest_full <- plot_anchor_forest(
  res_full$fisher,
  "Fisher's exact test: E2A-PBX1 peak enrichment\nper loop anchor class vs genome"
)
print(p_forest_full)
ggsave(file.path(loops_plots_dir, "loop_anchor_EP1_forest.pdf"),
       p_forest_full, width = 8, height = 5)

# ============================================================================
# B. Top25 697HF E2A-PBX1 peaks
# ============================================================================
cat("\n--- Top25 E2A-PBX1 peaks ---\n")
res_top25 <- compute_overlap_stats(anchors_gr_list, peaks_top25_gr, genome_file)

print(res_top25$fisher[, c("anchor_class", "n_anchors", "overlap_bp",
                            "pct_overlap", "fold_enrichment",
                            "OR", "CI_low", "CI_high", "p_value", "p_adj", "significance")])

write.table(res_top25$fisher,
            file.path(loops_files_dir, "loop_anchor_EP1_top25_chip_fisher.tsv"),
            sep = "\t", row.names = FALSE, quote = FALSE)

p_polar_top25 <- plot_anchor_polar(
  res_top25$counts,
  "Top25 E2A-PBX1 peak enrichment per loop anchor class\n(genome background)"
)
print(p_polar_top25)
ggsave(file.path(loops_plots_dir, "loop_anchor_EP1_top25_polar.pdf"),
       p_polar_top25, width = 10, height = 6)

p_forest_top25 <- plot_anchor_forest(
  res_top25$fisher,
  "Fisher's exact test: Top25 E2A-PBX1 peak enrichment\nper loop anchor class vs genome"
)
print(p_forest_top25)
ggsave(file.path(loops_plots_dir, "loop_anchor_EP1_top25_forest.pdf"),
       p_forest_top25, width = 8, height = 5)
```

```{r Double-anchor loops}
# ============================================================================
# Double-anchor analysis: loops with E2A-PBX1 at BOTH anchors
# ============================================================================
library(GenomicRanges)
library(ggplot2)
library(dplyr)
library(tidyr)

# ----------------------------------------------------------------------------
# 1. Load loop BEDPE files
# ----------------------------------------------------------------------------
loops_dir <- "R:/R_STIK/Hi-C/HiC_marta_Oct25/loops/intersected_loops"

loop_files <- list(
  "ARO_specific"   = "ARO_specific_only_05_10kb_exp30kb.bedpe",
  "EP1_specific"   = "EP1_specific_only_05_10kb_exp30kb.bedpe",
  "REH_specific"   = "REH_specific_only_05_10kb_exp30kb.bedpe",
  "shared_EP1_ARO" = "shared_EP1_ARO_only_05_10kb_exp30kb.bedpe",
  "shared_ARO_REH" = "shared_ARO_REH_only_05_10kb_exp30kb.bedpe",
  "shared_EP1_REH" = "shared_EP1_REH_only_05_10kb_exp30kb.bedpe",
  "shared_all3"    = "shared_loops_all3_05_10kb_exp30kb.bedpe"
)

read_loops <- function(filename) {
  df <- read.table(file.path(loops_dir, filename),
                   sep    = "\t",
                   header = FALSE)
  df %>%
    select(chr1 = 1, start1 = 2, end1 = 3,
           chr2 = 4, start2 = 5, end2 = 6) %>%
    mutate(start1  = as.integer(start1),
           end1    = as.integer(end1),
           start2  = as.integer(start2),
           end2    = as.integer(end2),
           loop_id = row_number())
}

loops_list <- lapply(loop_files, read_loops)

cat("Loop sets loaded:\n")
for (nm in names(loops_list)) {
  cat(sprintf("  %-20s : %d loops\n", nm, nrow(loops_list[[nm]])))
}

# ----------------------------------------------------------------------------
# 2. Helper: classify each loop as none / single / double anchor
# ----------------------------------------------------------------------------
classify_double_anchor <- function(loops_df, peaks_gr) {

  anc1_df <- data.frame(seqnames = loops_df$chr1,
                        start    = loops_df$start1,
                        end      = loops_df$end1,
                        loop_id  = loops_df$loop_id)
  anc2_df <- data.frame(seqnames = loops_df$chr2,
                        start    = loops_df$start2,
                        end      = loops_df$end2,
                        loop_id  = loops_df$loop_id)

  anc1_gr <- makeGRangesFromDataFrame(anc1_df, keep.extra.columns = TRUE)
  anc2_gr <- makeGRangesFromDataFrame(anc2_df, keep.extra.columns = TRUE)

  ids_anc1 <- unique(anc1_gr$loop_id[queryHits(findOverlaps(anc1_gr, peaks_gr))])
  ids_anc2 <- unique(anc2_gr$loop_id[queryHits(findOverlaps(anc2_gr, peaks_gr))])

  loops_df %>%
    mutate(peak_anc1    = loop_id %in% ids_anc1,
           peak_anc2    = loop_id %in% ids_anc2,
           anchor_class = case_when(
             peak_anc1 & peak_anc2 ~ "double",
             peak_anc1 | peak_anc2 ~ "single",
             TRUE                  ~ "none"
           ))
}

# ----------------------------------------------------------------------------
# 3. Run classification for both peak sets
# ----------------------------------------------------------------------------
classify_all <- function(loops_list, peaks_gr, peak_label) {

  results <- lapply(names(loops_list), function(nm) {
    df     <- classify_double_anchor(loops_list[[nm]], peaks_gr)
    counts <- dplyr::count(df, anchor_class) %>%
      dplyr::mutate(loop_class  = nm,
                    peak_set    = peak_label,
                    total_loops = nrow(df),
                    pct         = n / total_loops * 100)
    counts
  }) %>% dplyr::bind_rows()

  results %>%
    dplyr::mutate(
      loop_class   = factor(loop_class,   levels = anchor_levels),
      anchor_class = factor(anchor_class, levels = c("double", "single", "none"))
    )
}

res_double_full  <- classify_all(loops_list, peaks_gr,       "Full E2A-PBX1")
res_double_top25 <- classify_all(loops_list, peaks_top25_gr, "Top25 E2A-PBX1")

cat("\n--- Double-anchor summary: Full peaks ---\n")
print(res_double_full  %>% select(loop_class, anchor_class, n, pct, total_loops))
cat("\n--- Double-anchor summary: Top25 peaks ---\n")
print(res_double_top25 %>% select(loop_class, anchor_class, n, pct, total_loops))

write.table(res_double_full,
            file.path(loops_files_dir, "double_anchor_full_peaks.tsv"),
            sep = "\t", row.names = FALSE, quote = FALSE)
write.table(res_double_top25,
            file.path(loops_files_dir, "double_anchor_top25_peaks.tsv"),
            sep = "\t", row.names = FALSE, quote = FALSE)

# ----------------------------------------------------------------------------
# 4. Fisher test: double-anchor enrichment vs genome background (bp-level)
# ----------------------------------------------------------------------------
fisher_double_genome <- function(loops_list, peaks_gr, genome_file, peak_label) {

  # Genome size
  chrom_sizes <- read.table(genome_file, sep = "\t", header = FALSE,
                             col.names = c("chrom", "size")) %>%
    filter(chrom %in% seqlevels(peaks_gr))
  genome_size <- sum(as.numeric(chrom_sizes$size))

  # Peak bp coverage
  peaks_reduced <- reduce(peaks_gr)
  peak_bp       <- sum(as.numeric(width(peaks_reduced)))
  non_peak_bp   <- genome_size - peak_bp
  scale         <- 1e6 / genome_size

  c_bg <- round(peak_bp     * scale)
  d_bg <- round(non_peak_bp * scale)

  lapply(names(loops_list), function(nm) {

    loops_df <- loops_list[[nm]]
    df       <- classify_double_anchor(loops_df, peaks_gr)

    # For double-anchor loops: combine both anchors as one region set
    double_loops <- df %>% filter(anchor_class == "double")
    total_loops  <- nrow(df)
    n_double     <- nrow(double_loops)

    if (n_double == 0) {
      return(data.frame(loop_class  = nm,
                        peak_set    = peak_label,
                        n_double    = 0,
                        n_total     = total_loops,
                        pct_double  = 0,
                        overlap_bp  = 0,
                        OR          = NA,
                        CI_low      = NA,
                        p_value     = NA))
    }

    # bp of double-anchor loop anchors overlapping peaks
    anc1_double <- makeGRangesFromDataFrame(
      data.frame(seqnames = double_loops$chr1,
                 start    = double_loops$start1,
                 end      = double_loops$end1),
      seqnames.field = "seqnames"
    )
    anc2_double <- makeGRangesFromDataFrame(
      data.frame(seqnames = double_loops$chr2,
                 start    = double_loops$start2,
                 end      = double_loops$end2),
      seqnames.field = "seqnames"
    )

    both_anchors_gr  <- reduce(c(anc1_double, anc2_double))
    anchor_bp        <- sum(as.numeric(width(both_anchors_gr)))
    overlap_gr       <- intersect(both_anchors_gr, peaks_reduced)
    overlap_bp       <- sum(as.numeric(width(overlap_gr)))
    no_overlap_bp    <- anchor_bp - overlap_bp
    pct_overlap      <- overlap_bp / anchor_bp * 100

    ct <- matrix(c(round(overlap_bp), round(no_overlap_bp),
                   c_bg,              d_bg),
                 nrow = 2, byrow = TRUE,
                 dimnames = list(c("Double anchors", "Genome"),
                                 c("Peak_bp", "Non_peak_bp")))

    ft <- fisher.test(ct, alternative = "greater")

    data.frame(loop_class  = nm,
               peak_set    = peak_label,
               n_double    = n_double,
               n_total     = total_loops,
               pct_double  = n_double / total_loops * 100,
               anchor_bp   = anchor_bp,
               overlap_bp  = overlap_bp,
               pct_overlap = pct_overlap,
               OR          = ft$estimate,
               CI_low      = ft$conf.int[1],
               CI_high     = ft$conf.int[2],
               p_value     = ft$p.value)
  }) %>% bind_rows() %>%
    mutate(
      p_adj        = p.adjust(p_value, method = "BH"),
      significance = case_when(
        p_adj < 0.001 ~ "***",
        p_adj < 0.01  ~ "**",
        p_adj < 0.05  ~ "*",
        TRUE          ~ "ns"
      ),
      loop_class = factor(loop_class, levels = anchor_levels)
    )
}

fisher_double_full  <- fisher_double_genome(loops_list, peaks_gr,       genome_file, "Full E2A-PBX1")
fisher_double_top25 <- fisher_double_genome(loops_list, peaks_top25_gr, genome_file, "Top25 E2A-PBX1")

cat("\n--- Fisher double-anchor vs genome: Full peaks ---\n")
print(fisher_double_full)
cat("\n--- Fisher double-anchor vs genome: Top25 peaks ---\n")
print(fisher_double_top25)

write.table(dplyr::bind_rows(fisher_double_full, fisher_double_top25),
            file.path(loops_files_dir, "double_anchor_fisher_genome.tsv"),
            sep = "\t", row.names = FALSE, quote = FALSE)

# ----------------------------------------------------------------------------
# 5. Stacked barplot: double / single / none proportions
# ----------------------------------------------------------------------------
double_colors <- c("double" = "#c0392b",
                   "single" = "#e67e22",
                   "none"   = "#bdc3c7")

plot_double_bar <- function(res_df, title_str) {
  ggplot(res_df, aes(x = loop_class, y = pct, fill = anchor_class)) +
    geom_bar(stat = "identity", color = "black", linewidth = 0.4) +
    geom_text(aes(label = ifelse(pct > 2, paste0(round(pct, 1), "%"), "")),
              position = position_stack(vjust = 0.5),
              size = 3.5, fontface = "bold", color = "white") +
    scale_fill_manual(values = double_colors,
                      name   = "Peak overlap",
                      labels = c("double" = "Both anchors",
                                 "single" = "One anchor",
                                 "none"   = "No anchor")) +
    scale_y_continuous(expand = c(0, 0), limits = c(0, 101)) +
    theme_bw() +
    theme(
      panel.grid   = element_blank(),
      axis.text.x  = element_text(size = 11, angle = 35, hjust = 1, face = "bold"),
      axis.text.y  = element_text(size = 10),
      axis.title   = element_text(size = 13),
      plot.title   = element_text(size = 14, face = "bold", hjust = 0.5),
      legend.title = element_text(face = "bold")
    ) +
    xlab("Loop class") + ylab("Proportion of loops (%)") +
    labs(title = title_str)
}

p_double_full <- plot_double_bar(res_double_full,
                                  "E2A-PBX1 peak overlap at loop anchors\n(Full peaks)")
print(p_double_full)
ggsave(file.path(loops_plots_dir, "double_anchor_full_bar.pdf"),
       p_double_full, width = 10, height = 6)

p_double_top25 <- plot_double_bar(res_double_top25,
                                   "E2A-PBX1 peak overlap at loop anchors\n(Top25 peaks)")
print(p_double_top25)
ggsave(file.path(loops_plots_dir, "double_anchor_top25_bar.pdf"),
       p_double_top25, width = 10, height = 6)

# ----------------------------------------------------------------------------
# 6. Dot plot: % double-anchor with genome-background significance
# ----------------------------------------------------------------------------
plot_double_dot <- function(fisher_df, title_str) {
  ggplot(fisher_df, aes(x = loop_class, y = pct_double, fill = loop_class)) +
    geom_col(color = "black", linewidth = 0.4, width = 0.7) +
    geom_text(aes(label = significance, y = pct_double),
              vjust = -0.4, size = 5, color = "black") +
    scale_fill_manual(values = anchor_colors) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
    theme_bw() +
    theme(
      panel.grid      = element_blank(),
      axis.text.x     = element_text(size = 11, angle = 35, hjust = 1, face = "bold"),
      axis.text.y     = element_text(size = 10),
      axis.title      = element_text(size = 13),
      plot.title      = element_text(size = 14, face = "bold", hjust = 0.5),
      legend.position = "none"
    ) +
    xlab("Loop class") +
    ylab("% loops with peak at both anchors") +
    labs(title = title_str)
}

p_dot_full <- plot_double_dot(
  fisher_double_full,
  "Loops with E2A-PBX1 at both anchors\n(Full peaks, vs genome background)"
)
print(p_dot_full)
ggsave(file.path(loops_plots_dir, "double_anchor_full_dot.pdf"),
       p_dot_full, width = 10, height = 6)

p_dot_top25 <- plot_double_dot(
  fisher_double_top25,
  "Loops with E2A-PBX1 at both anchors\n(Top25 peaks, vs genome background)"
)
print(p_dot_top25)
ggsave(file.path(loops_plots_dir, "double_anchor_top25_dot.pdf"),
       p_dot_top25, width = 10, height = 6)
```

---
# Session info

```{r session-info}
sessionInfo()
```
