---
title: "Hi-C TAD Analysis – REH samples"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: hide
---
# Overview

This notebook analyses **TAD boundaries** inferred from Hi-C data using
HiCExplorer findTADs bed files and posterior bedtools comparisons, obtaining three boundary conditions:

| Condition | Description |
|-----------|-------------|
| **REH** | Specific boudndaries parental cell line |
| **EP1** | Specific boundaries EP1 treatment |
| **shared** | shared boundaries | 

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)
knitr::opts_knit$set(root.dir = "R:/R_STIK/Hi-C/HiC_marta_Oct25/TADs/comparison")
# knitr::opts_knit$set(root.dir = "/ijc/LABS/STIK/RAW/Hi-C/HiC_marta_Oct25/TADs/comparison")

```

First we are going to plot the coverage of REH ATAC-Seq reads into the different boundary conditions, and also with a randomized regions negative control.

```{r ATAC-Seq intersection, eval=FALSE}
library(dplyr)
library(ggplot2)
library(ggpubr)
library(rstatix)
library(tidyr)
library(patchwork)

# ============================================================================
# Load data
# ============================================================================
total_reads <- as.numeric(readLines(
  "R:/R_STIK/Hi-C/HiC_marta_Oct25/TADs/comparison/atac_coverage/total_mapped_reads.txt"
))

setwd("R:/R_STIK/Hi-C/HiC_marta_Oct25/TADs/comparison/atac_coverage")

col_names <- c("chr", "start", "end", "boundary_class",
               "read_count", "bases_covered",
               "boundary_length", "fraction_covered")

cov <- read.table("boundary_coverage_full.bed",
                  sep = "\t", col.names = col_names)

rnd <- read.table("random_coverage_full.bed",
                  sep = "\t", col.names = col_names) %>%
  mutate(boundary_class = "random")

# ============================================================================
# Level order & colors
# ============================================================================
boundary_class_levels <- c("shared", "unique_HiC_REH_EP1", "unique_REH")
level_order_full      <- c(boundary_class_levels, "random")

class_colors <- c(
  "shared"             = "#3498db",
  "unique_HiC_REH_EP1" = "#9b59b6",
  "unique_REH"         = "#e74c3c",
  "random"             = "#95a5a6"
)

x_labels <- c(
  "shared"             = "Shared",
  "unique_HiC_REH_EP1" = "Unique HiC-REH-EP1",
  "unique_REH"         = "Unique REH",
  "random"             = "Random"
)

# ============================================================================
# Compute metrics — boundaries only (pairwise)
# ============================================================================
cov_boundaries <- cov %>%
  mutate(
    CPM               = (read_count / total_reads) * 1e6,
    log2_CPM          = log2(CPM + 1),
    mean_depth        = read_count / boundary_length,
    mean_depth_pseudo = mean_depth + 1e-6,
    boundary_class    = factor(boundary_class, levels = boundary_class_levels)
  )

# ============================================================================
# Compute metrics — all classes including random (vs-random)
# ============================================================================
all_cov <- bind_rows(cov, rnd) %>%
  mutate(
    CPM               = (read_count / total_reads) * 1e6,
    log2_CPM          = log2(CPM + 1),
    mean_depth        = read_count / boundary_length,
    mean_depth_pseudo = mean_depth + 1e-6,
    boundary_class    = factor(boundary_class, levels = level_order_full)
  )

# ============================================================================
# Helper: build bracket segments + labels from sig_ann data frame
# ============================================================================
make_brackets <- function(sig_ann, tick_drop = 0.05, label_lift = 0.07) {
  segments <- bind_rows(
    sig_ann %>% transmute(x = x,    xend = xend, y = y.position,             yend = y.position),
    sig_ann %>% transmute(x = x,    xend = x,    y = y.position - tick_drop, yend = y.position),
    sig_ann %>% transmute(x = xend, xend = xend, y = y.position - tick_drop, yend = y.position)
  )
  labels <- sig_ann %>%
    transmute(x = xmid, y = y.position + label_lift, label = p.adj.signif)
  list(segments = segments, labels = labels)
}

# ============================================================================
# PART 1 — Pairwise Wilcoxon between boundary classes (no random)
# ============================================================================
wilcox_pairwise <- cov_boundaries %>%
  wilcox_test(log2_CPM ~ boundary_class) %>%
  adjust_pvalue(method = "BH") %>%
  add_significance()

sig_pairwise <- wilcox_pairwise %>%
  filter(p.adj < 0.05) %>%
  select(group1, group2, p.adj.signif)

x_pos_boundaries <- setNames(seq_along(boundary_class_levels), boundary_class_levels)

y_max_b   <- max(cov_boundaries$log2_CPM, na.rm = TRUE)
y_start_b <- y_max_b * 0.75

if (nrow(sig_pairwise) > 0) {
  sig_pairwise <- sig_pairwise %>%
    mutate(
      y.position = seq(y_start_b,
                       y_start_b + (y_max_b * 0.22),
                       length.out = n()),
      x    = x_pos_boundaries[group1],
      xend = x_pos_boundaries[group2],
      xmid = (x_pos_boundaries[group1] + x_pos_boundaries[group2]) / 2
    )
  brk_pairwise <- make_brackets(sig_pairwise)
} else {
  brk_pairwise <- list(
    segments = data.frame(x = numeric(), xend = numeric(), y = numeric(), yend = numeric()),
    labels   = data.frame(x = numeric(), y = numeric(), label = character())
  )
}

# --- Plot A1: boxplot, pairwise ---
p_box_pairwise <- ggplot(cov_boundaries, aes(x = boundary_class, y = log2_CPM, fill = boundary_class)) +
  geom_boxplot(outlier.size = 0.3, outlier.alpha = 0.4, size = 0.4, width = 0.65) +
  geom_hline(
    yintercept = median(cov_boundaries$log2_CPM, na.rm = TRUE),
    linetype = "dashed", color = "red", size = 0.5
  ) +
  geom_segment(
    data = brk_pairwise$segments,
    aes(x = x, xend = xend, y = y, yend = yend),
    inherit.aes = FALSE, size = 0.4
  ) +
  geom_text(
    data = brk_pairwise$labels,
    aes(x = x, y = y, label = label),
    inherit.aes = FALSE, size = 4
  ) +
  scale_fill_manual(values = class_colors) +
  scale_x_discrete(labels = x_labels) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.22))) +
  theme_bw() +
  theme(
    panel.grid      = element_blank(),
    axis.text.x     = element_text(angle = 35, hjust = 1, size = 10, face = "bold"),
    axis.text.y     = element_text(size = 10),
    axis.title      = element_text(size = 12),
    plot.title      = element_text(hjust = 0.5, face = "bold", size = 13),
    plot.subtitle   = element_text(hjust = 0.5, size = 9, color = "gray40"),
    legend.position = "none"
  ) +
  xlab("Boundary type") + ylab("Coverage (log2 CPM)") +
  labs(
    title    = "REH ATAC-seq signal at TAD boundaries",
    subtitle = "Pairwise Wilcoxon, BH-adjusted — significant comparisons only"
  )

# --- Plot B1: violin + boxplot, pairwise ---
p_violin_pairwise <- ggplot(cov_boundaries, aes(x = boundary_class, y = mean_depth_pseudo, fill = boundary_class)) +
  geom_violin(alpha = 0.65, scale = "width", trim = FALSE, size = 0.3) +
  geom_boxplot(width = 0.12, outlier.size = 0.3, outlier.alpha = 0.4,
               size = 0.4, fill = "white", color = "black") +
  geom_hline(
    yintercept = median(cov_boundaries$mean_depth_pseudo, na.rm = TRUE),
    linetype = "dashed", color = "red", size = 0.5
  ) +
  scale_y_log10(
    breaks = c(0.001, 0.01, 0.1, 1),
    labels = c("0.001", "0.010", "0.100", "1.000")
  ) +
  scale_fill_manual(values = class_colors) +
  scale_x_discrete(labels = x_labels) +
  theme_bw() +
  theme(
    panel.grid      = element_blank(),
    axis.text.x     = element_text(angle = 35, hjust = 1, size = 10, face = "bold"),
    axis.text.y     = element_text(size = 10),
    axis.title      = element_text(size = 12),
    plot.title      = element_text(hjust = 0.5, face = "bold", size = 13),
    legend.position = "none"
  ) +
  xlab("Boundary type") + ylab("Mean depth (log10 scale)") +
  labs(title = "REH ATAC-seq coverage at TAD boundaries")

# --- Plot C1: fraction covered, pairwise ---
p_fraction_pairwise <- ggplot(cov_boundaries, aes(x = boundary_class, y = fraction_covered, fill = boundary_class)) +
  geom_boxplot(outlier.size = 0.3, outlier.alpha = 0.4, size = 0.4, width = 0.65) +
  geom_hline(
    yintercept = median(cov_boundaries$fraction_covered, na.rm = TRUE),
    linetype = "dashed", color = "red", size = 0.5
  ) +
  scale_fill_manual(values = class_colors) +
  scale_x_discrete(labels = x_labels) +
  scale_y_continuous(limits = c(0, 1), expand = expansion(mult = c(0.02, 0.05))) +
  theme_bw() +
  theme(
    panel.grid      = element_blank(),
    axis.text.x     = element_text(angle = 35, hjust = 1, size = 10, face = "bold"),
    axis.text.y     = element_text(size = 10),
    axis.title      = element_text(size = 12),
    plot.title      = element_text(hjust = 0.5, face = "bold", size = 13),
    legend.position = "none"
  ) +
  xlab("Boundary type") + ylab("Fraction of bases covered") +
  labs(title = "ATAC-seq breadth of coverage at TAD boundaries")

# ============================================================================
# PART 2 — Wilcoxon: each boundary class vs random, BH correction
# ============================================================================

# --- log2 CPM vs random ---
wilcox_vs_random <- lapply(boundary_class_levels, function(cls) {
  sub_df <- all_cov %>%
    filter(boundary_class %in% c(cls, "random")) %>%
    mutate(boundary_class = droplevels(boundary_class))
  
  wilcox_test(sub_df, log2_CPM ~ boundary_class, ref.group = "random") %>%
    mutate(group1 = cls, group2 = "random")
}) %>%
  bind_rows() %>%
  adjust_pvalue(method = "BH") %>%
  add_significance()

sig_vs_random <- wilcox_vs_random %>%
  filter(p.adj < 0.05) %>%
  select(group1, group2, p.adj.signif)

x_pos_full   <- setNames(seq_along(level_order_full), level_order_full)
y_max_full   <- max(all_cov$log2_CPM, na.rm = TRUE)
y_start_full <- y_max_full * 0.80

if (nrow(sig_vs_random) > 0) {
  sig_vs_random <- sig_vs_random %>%
    mutate(
      y.position = seq(y_start_full,
                       y_start_full + (y_max_full * 0.18),
                       length.out = n()),
      x    = x_pos_full[group1],
      xend = x_pos_full[group2],
      xmid = (x_pos_full[group1] + x_pos_full[group2]) / 2
    )
  brk_random <- make_brackets(sig_vs_random)
} else {
  brk_random <- list(
    segments = data.frame(x = numeric(), xend = numeric(), y = numeric(), yend = numeric()),
    labels   = data.frame(x = numeric(), y = numeric(), label = character())
  )
}

# --- fraction covered vs random ---
wilcox_frac_vs_random <- lapply(boundary_class_levels, function(cls) {
  sub_df <- all_cov %>%
    filter(boundary_class %in% c(cls, "random")) %>%
    mutate(boundary_class = droplevels(boundary_class))
  
  wilcox_test(sub_df, fraction_covered ~ boundary_class, ref.group = "random") %>%
    mutate(group1 = cls, group2 = "random")
}) %>%
  bind_rows() %>%
  adjust_pvalue(method = "BH") %>%
  add_significance()

sig_frac_random <- wilcox_frac_vs_random %>%
  filter(p.adj < 0.05) %>%
  select(group1, group2, p.adj.signif)

if (nrow(sig_frac_random) > 0) {
  sig_frac_random <- sig_frac_random %>%
    mutate(
      y.position = seq(0.80, 0.95, length.out = n()),
      x    = x_pos_full[group1],
      xend = x_pos_full[group2],
      xmid = (x_pos_full[group1] + x_pos_full[group2]) / 2
    )
  brk_frac_random <- make_brackets(sig_frac_random, tick_drop = 0.02, label_lift = 0.03)
} else {
  brk_frac_random <- list(
    segments = data.frame(x = numeric(), xend = numeric(), y = numeric(), yend = numeric()),
    labels   = data.frame(x = numeric(), y = numeric(), label = character())
  )
}

# --- Plot A2: boxplot, vs random ---
p_box_random <- ggplot(all_cov, aes(x = boundary_class, y = log2_CPM, fill = boundary_class)) +
  geom_boxplot(outlier.size = 0.3, outlier.alpha = 0.4, size = 0.4, width = 0.65) +
  geom_hline(
    yintercept = median(all_cov$log2_CPM[all_cov$boundary_class == "random"], na.rm = TRUE),
    linetype = "dashed", color = "gray40", size = 0.5
  ) +
  geom_segment(
    data = brk_random$segments,
    aes(x = x, xend = xend, y = y, yend = yend),
    inherit.aes = FALSE, size = 0.4
  ) +
  geom_text(
    data = brk_random$labels,
    aes(x = x, y = y, label = label),
    inherit.aes = FALSE, size = 4
  ) +
  scale_fill_manual(values = class_colors) +
  scale_x_discrete(labels = x_labels) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.22))) +
  theme_bw() +
  theme(
    panel.grid      = element_blank(),
    axis.text.x     = element_text(angle = 35, hjust = 1, size = 10, face = "bold"),
    axis.text.y     = element_text(size = 10),
    axis.title      = element_text(size = 12),
    plot.title      = element_text(hjust = 0.5, face = "bold", size = 13),
    plot.subtitle   = element_text(hjust = 0.5, size = 9, color = "gray40"),
    legend.position = "none"
  ) +
  xlab("Boundary type") + ylab("Coverage (log2 CPM)") +
  labs(
    title    = "REH ATAC-seq signal at TAD boundaries",
    subtitle = "Wilcoxon vs random, BH-adjusted — significant comparisons only"
  )

# --- Plot B2: violin + boxplot, vs random ---
p_violin_random <- ggplot(all_cov, aes(x = boundary_class, y = mean_depth_pseudo, fill = boundary_class)) +
  geom_violin(alpha = 0.65, scale = "width", trim = FALSE, size = 0.3) +
  geom_boxplot(width = 0.12, outlier.size = 0.3, outlier.alpha = 0.4,
               size = 0.4, fill = "white", color = "black") +
  geom_hline(
    yintercept = median(all_cov$mean_depth_pseudo[all_cov$boundary_class == "random"], na.rm = TRUE),
    linetype = "dashed", color = "gray40", size = 0.5
  ) +
  scale_y_log10(
    breaks = c(0.001, 0.01, 0.1, 1),
    labels = c("0.001", "0.010", "0.100", "1.000")
  ) +
  scale_fill_manual(values = class_colors) +
  scale_x_discrete(labels = x_labels) +
  theme_bw() +
  theme(
    panel.grid      = element_blank(),
    axis.text.x     = element_text(angle = 35, hjust = 1, size = 10, face = "bold"),
    axis.text.y     = element_text(size = 10),
    axis.title      = element_text(size = 12),
    plot.title      = element_text(hjust = 0.5, face = "bold", size = 13),
    legend.position = "none"
  ) +
  xlab("Boundary type") + ylab("Mean depth (log10 scale)") +
  labs(title = "REH ATAC-seq coverage at TAD boundaries (+ random)")

# --- Plot C2: fraction covered, vs random ---
p_fraction_random <- ggplot(all_cov, aes(x = boundary_class, y = fraction_covered, fill = boundary_class)) +
  geom_boxplot(outlier.size = 0.3, outlier.alpha = 0.4, size = 0.4, width = 0.65) +
  geom_hline(
    yintercept = median(all_cov$fraction_covered[all_cov$boundary_class == "random"], na.rm = TRUE),
    linetype = "dashed", color = "gray40", size = 0.5
  ) +
  geom_segment(
    data = brk_frac_random$segments,
    aes(x = x, xend = xend, y = y, yend = yend),
    inherit.aes = FALSE, size = 0.4
  ) +
  geom_text(
    data = brk_frac_random$labels,
    aes(x = x, y = y, label = label),
    inherit.aes = FALSE, size = 4
  ) +
  scale_fill_manual(values = class_colors) +
  scale_x_discrete(labels = x_labels) +
  scale_y_continuous(limits = c(0, 1.1), expand = expansion(mult = c(0.02, 0.05))) +
  theme_bw() +
  theme(
    panel.grid      = element_blank(),
    axis.text.x     = element_text(angle = 35, hjust = 1, size = 10, face = "bold"),
    axis.text.y     = element_text(size = 10),
    axis.title      = element_text(size = 12),
    plot.title      = element_text(hjust = 0.5, face = "bold", size = 13),
    plot.subtitle   = element_text(hjust = 0.5, size = 9, color = "gray40"),
    legend.position = "none"
  ) +
  xlab("Boundary type") + ylab("Fraction of bases covered") +
  labs(
    title    = "ATAC-seq breadth of coverage at TAD boundaries (+ random)",
    subtitle = "Wilcoxon vs random, BH-adjusted — significant comparisons only"
  )
print(p_combined_pairwise)
print(p_combined_random)
# ============================================================================
# Save — pairwise (boundary classes only)
# ============================================================================
pdf("REH_atac_boundary_boxplot.pdf", width = 6, height = 6)
print(p_box_pairwise)
dev.off()

pdf("REH_atac_boundary_violin.pdf", width = 6, height = 6)
print(p_violin_pairwise)
dev.off()

pdf("REH_atac_boundary_fraction.pdf", width = 6, height = 6)
print(p_fraction_pairwise)
dev.off()

p_combined_pairwise <- p_box_pairwise + p_violin_pairwise + p_fraction_pairwise +
  plot_annotation(tag_levels = "A")

pdf("REH_atac_boundary_combined.pdf", width = 18, height = 6)
print(p_combined_pairwise)
dev.off()

# ============================================================================
# Save — vs random
# ============================================================================
pdf("REH_atac_boundary+random_boxplot.pdf", width = 7, height = 6)
print(p_box_random)
dev.off()

pdf("REH_atac_boundary+random_violin.pdf", width = 7, height = 6)
print(p_violin_random)
dev.off()

pdf("REH_atac_boundary+random_fraction.pdf", width = 7, height = 6)
print(p_fraction_random)
dev.off()

p_combined_random <- p_box_random + p_violin_random + p_fraction_random +
  plot_annotation(tag_levels = "A")

pdf("REH_atac_boundary+random_combined.pdf", width = 21, height = 6)
print(p_combined_random)
dev.off()

# ============================================================================
# Summary tables & stats printout
# ============================================================================
summary_tbl <- all_cov %>%
  group_by(boundary_class) %>%
  summarise(
    n_regions       = n(),
    median_log2_CPM = median(log2_CPM, na.rm = TRUE),
    mean_log2_CPM   = mean(log2_CPM, na.rm = TRUE),
    median_depth    = median(mean_depth, na.rm = TRUE),
    median_frac_cov = median(fraction_covered, na.rm = TRUE),
    .groups = "drop"
  )

print(summary_tbl)
write.table(summary_tbl, "per_class_summary_R.tsv",
            sep = "\t", quote = FALSE, row.names = FALSE)

cat("\n--- Wilcoxon pairwise results (boundary classes only) ---\n")
print(wilcox_pairwise %>% select(group1, group2, p, p.adj, p.adj.signif))

cat("\n--- Wilcoxon vs random results (log2 CPM) ---\n")
print(wilcox_vs_random %>% select(group1, group2, p, p.adj, p.adj.signif))

cat("\n--- Wilcoxon vs random results (fraction covered) ---\n")
print(wilcox_frac_vs_random %>% select(group1, group2, p, p.adj, p.adj.signif))
```
We are also going to validate the TADs using CTCF ChIP public data for REH cells.

```{r CTCF ChIP-Seq intersection}

```