---
title: "TADs_RNA-Seq"
author: "Xavier Vilchez-Perez"
date: "2026-03-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
# knitr::opts_knit$set(root.dir = "/ijc/LABS/STIK/RAW/Hi-C/Xavi_TADs/RCH-iMac") ### OnDemand server
knitr::opts_knit$set(root.dir = "R:/R_STIK/Hi-C/Xavi_TADs/datasets_comparison") ### Workstation
```

## Overview

This document processes RNA-seq count matrices from three datasets into comparable
**log2-centered expression matrices**. All three datasets follow the same normalization
strategy to ensure comparability:

| Dataset | Raw input | Normalization | Transformation |
|---------|-----------|---------------|----------------|
| BLaER   | Pre-processed BED | DESeq2 median-of-ratios (`normalized=TRUE`) | log2 → row-centered (as-is) |
| RCH     | Raw counts (.tab) | DESeq2 median-of-ratios (`counts(dds, normalized=TRUE)`) | log2(x+1) → row-center |
| 293     | Raw counts (.txt) | DESeq2 median-of-ratios (`counts(dds, normalized=TRUE)`) | log2(x+1) → row-center |

All datasets are annotated with **HGNC gene symbols** and deduplicated by keeping the
highest-expressed entry per symbol.

---

## Load Libraries

```{r libraries}
library(tidyverse)
library(DESeq2)
library(org.Hs.eg.db)
library(AnnotationDbi)
```

---

## Helper function: Ensembl → gene symbol annotation + deduplication

```{r symbol_helper}
annotate_and_deduplicate <- function(df, numeric_cols) {
  # df must have version-stripped Ensembl IDs as rownames
  # numeric_cols: character vector of column names with expression values

  ensembl_ids <- rownames(df)

  # Map Ensembl IDs to HGNC gene symbols
  symbols <- mapIds(
    org.Hs.eg.db,
    keys      = ensembl_ids,
    column    = "SYMBOL",
    keytype   = "ENSEMBL",
    multiVals = "first"
  )

  valid <- !is.na(symbols)
  cat(paste("  Total genes:               ", length(symbols), "\n"))
  cat(paste("  Genes with valid symbols:  ", sum(valid), "\n"))
  cat(paste("  Genes removed (no symbol): ", sum(!valid), "\n\n"))

  # Filter to genes with valid symbols
  df_sym <- df[valid, , drop = FALSE]
  df_sym$symbol <- symbols[valid]
  rownames(df_sym) <- NULL

  # Deduplicate: keep the row with highest mean expression per symbol
  df_sym$mean_expr <- rowMeans(df_sym[, numeric_cols, drop = FALSE])
  df_sym <- df_sym %>%
    group_by(symbol) %>%
    slice_max(mean_expr, n = 1, with_ties = FALSE) %>%
    ungroup()

  cat(paste("  Final unique genes after deduplication:", nrow(df_sym), "\n\n"))

  # Restore rownames and return only expression columns
  df_sym$mean_expr <- NULL
  df_sym <- as.data.frame(df_sym)
  rownames(df_sym) <- df_sym$symbol
  df_sym$symbol <- NULL

  return(df_sym[, numeric_cols, drop = FALSE])
}
```

---

## 1. BLaER Dataset

### Load and clean

The BED file already contains log2-centered values derived from DESeq2
`normalized=TRUE` (median-of-ratios). We extract the expression columns and retain
only the **first timepoint (0h → Ctrl)** and the **last timepoint (192h)**.

Columns 9–19 correspond to: 0, 9, 18, 24, 48, 72, 96, 120, 144, 168, 192 h.

```{r blaer_load}
# blaer_path <- "/ijc/LABS/STIK/RAW/Hi-C/Xavi_TADs/RCH-iMac/RNA_BLaER_log2centered_filter_bc_tss.bed"
blaer_path <- "R:/R_STIK/Hi-C/Xavi_TADs/RCH-iMac/RNA_BLaER_log2centered_filter_bc_tss.bed"

blaer_raw <- read.table(
  blaer_path,
  header       = FALSE,
  sep          = "\t",
  quote        = "",
  comment.char = "",
  stringsAsFactors = FALSE
)

# Cols 1-8: genomic/annotation info; cols 9-19: expression timepoints
timepoints <- c("0h", "9h", "18h", "24h", "48h", "72h", "96h", "120h", "144h", "168h", "192h")
colnames(blaer_raw) <- c("chr", "start", "end", "source", "feature",
                          "gene_id", "gene_type", "gene_name", timepoints)

# Strip Ensembl version suffix
blaer_raw$gene_id_clean <- sub("\\..*", "", blaer_raw$gene_id)
```

### Select Ctrl (0h) and 192h, deduplicate at Ensembl level

```{r blaer_select}
blaer_sub <- blaer_raw[, c("gene_id_clean", "0h", "192h")]
colnames(blaer_sub) <- c("gene_id", "Ctrl", "192h")

# Deduplicate Ensembl IDs before symbol mapping (keep highest mean expression)
blaer_sub$mean_expr <- rowMeans(blaer_sub[, c("Ctrl", "192h")])
blaer_sub <- blaer_sub %>%
  group_by(gene_id) %>%
  slice_max(mean_expr, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  as.data.frame()

blaer_sub$mean_expr <- NULL
rownames(blaer_sub) <- blaer_sub$gene_id
blaer_sub$gene_id   <- NULL
```

### Annotate with gene symbols

```{r blaer_symbols}
cat("=== BLaER symbol annotation ===\n")
blaer_log2centered <- annotate_and_deduplicate(blaer_sub, numeric_cols = c("Ctrl", "192h"))

head(blaer_log2centered)
dim(blaer_log2centered)
```

> **Note:** BLaER values are already log2-centered (DESeq2 median-of-ratios normalized).
> No further transformation applied — subset, deduplicate, and annotate only.

---

## 2. RCH Dataset

### Load raw counts

```{r rch_load}
# rch_path <- "/ijc/LABS/STIK/RAW/rna_seq/rnaseq_martaRCHsh_gen25/r_analysis/final_counts_RCH.tab"
rch_path <- "R:/R_STIK/rna_seq/rnaseq_martaRCHsh_gen25/r_analysis/final_counts_RCH.tab"

rch_raw <- read.table(
  rch_path,
  header       = FALSE,
  sep          = "\t",
  row.names    = 1,
  quote        = "",
  comment.char = "",
  stringsAsFactors = FALSE
)

colnames(rch_raw) <- c("Scr_1", "Scr_2", "PBX1_1", "PBX1_2")

# Remove HTSeq summary rows (N_unmapped, N_multimapping, etc.)
rch_counts <- rch_raw[!grepl("^N_", rownames(rch_raw)), ]

# Strip Ensembl version suffix
# Rare duplicates (same gene, different version) resolved by keeping highest expressed
rch_counts$gene_id  <- sub("\\..*", "", rownames(rch_counts))
rch_counts$mean_expr <- rowMeans(rch_counts[, c("Scr_1", "Scr_2", "PBX1_1", "PBX1_2")])
rch_counts <- rch_counts %>%
  group_by(gene_id) %>%
  slice_max(mean_expr, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  as.data.frame()
rch_counts$mean_expr <- NULL
rownames(rch_counts)  <- rch_counts$gene_id
rch_counts$gene_id    <- NULL

# DESeq2 requires a plain integer matrix
rch_counts <- as.matrix(rch_counts)
mode(rch_counts) <- "integer"

head(rch_counts)
dim(rch_counts)
```

### DESeq2 size-factor normalization (all 4 replicates together)

```{r rch_deseq2}
# Build colData
rch_coldata <- data.frame(
  condition = factor(c("shScr", "shScr", "shPBX1", "shPBX1")),
  row.names = colnames(rch_counts)
)

# Create DESeqDataSet and estimate size factors
rch_dds <- DESeqDataSetFromMatrix(
  countData = rch_counts,
  colData   = rch_coldata,
  design    = ~ condition
)

# Filter lowly expressed genes (total counts across all samples > 5)
rch_dds <- rch_dds[rowSums(counts(rch_dds)) > 5, ]
cat(paste("RCH genes kept after low-count filtering:", nrow(rch_dds), "\n\n"))

# Estimate size factors (median-of-ratios, same method as BLaER)
rch_dds <- estimateSizeFactors(rch_dds)
cat("RCH size factors:\n")
print(sizeFactors(rch_dds))

# Extract normalized counts (equivalent to DESeq2 normalized=TRUE)
rch_norm <- counts(rch_dds, normalized = TRUE)
```

### Merge replicates by mean → shScr and shPBX1

```{r rch_merge}
rch_merged <- data.frame(
  shScr  = rowMeans(rch_norm[, c("Scr_1",  "Scr_2")]),
  shPBX1 = rowMeans(rch_norm[, c("PBX1_1", "PBX1_2")]),
  row.names = rownames(rch_norm)
)
```

### log2(x+1) transform and row-center

```{r rch_log2center}
# log2(normalized_count + 1) — mirrors BLaER preprocessing
rch_log2 <- log2(rch_merged + 1)

# Row-center: subtract each gene's mean across conditions
rch_row_means <- rowMeans(rch_log2)
rch_log2c     <- sweep(rch_log2, 1, rch_row_means, FUN = "-")
```

### Annotate with gene symbols

```{r rch_symbols}
cat("=== RCH symbol annotation ===\n")
rch_log2centered <- annotate_and_deduplicate(rch_log2c, numeric_cols = c("shScr", "shPBX1"))

head(rch_log2centered)
dim(rch_log2centered)
```

---

## 3. 293 Dataset

### Load raw counts

```{r counts293_load}
# counts293_path <- "/ijc/LABS/STIK/RAW/rna_seq/RNA_seq_23_June/analysis/rna_seq_june_23/data/final_counts_293.txt"
counts293_path <- "R:/R_STIK/rna_seq/RNA_seq_23_June/analysis/rna_seq_june_23/data/final_counts_293.txt"

counts293_raw <- read.table(
  counts293_path,
  header       = FALSE,
  sep          = "\t",
  row.names    = 1,
  quote        = "",
  comment.char = "",
  stringsAsFactors = FALSE
)

colnames(counts293_raw) <- c("EP_1", "EP_2", "NT_1", "NT_2")

# Remove HTSeq summary rows
counts293 <- counts293_raw[!grepl("^N_", rownames(counts293_raw)), ]

# Strip Ensembl version suffix
# Rare duplicates (same gene, different version) resolved by keeping highest expressed
counts293$gene_id   <- sub("\\..*", "", rownames(counts293))
counts293$mean_expr <- rowMeans(counts293[, c("EP_1", "EP_2", "NT_1", "NT_2")])
counts293 <- counts293 %>%
  group_by(gene_id) %>%
  slice_max(mean_expr, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  as.data.frame()
counts293$mean_expr <- NULL
rownames(counts293)  <- counts293$gene_id
counts293$gene_id    <- NULL

# DESeq2 requires a plain integer matrix
counts293 <- as.matrix(counts293)
mode(counts293) <- "integer"

head(counts293)
dim(counts293)
```

### DESeq2 size-factor normalization (all 4 replicates together)

```{r counts293_deseq2}
# Build colData
counts293_coldata <- data.frame(
  condition = factor(c("EP-GFP", "EP-GFP", "NT", "NT")),
  row.names = colnames(counts293)
)

# Create DESeqDataSet and estimate size factors
counts293_dds <- DESeqDataSetFromMatrix(
  countData = counts293,
  colData   = counts293_coldata,
  design    = ~ condition
)

# Filter lowly expressed genes (total counts across all samples > 5)
counts293_dds <- counts293_dds[rowSums(counts(counts293_dds)) > 5, ]
cat(paste("293 genes kept after low-count filtering:", nrow(counts293_dds), "\n\n"))

# Estimate size factors (median-of-ratios, same method as BLaER)
counts293_dds <- estimateSizeFactors(counts293_dds)
cat("293 size factors:\n")
print(sizeFactors(counts293_dds))

# Extract normalized counts (equivalent to DESeq2 normalized=TRUE)
counts293_norm <- counts(counts293_dds, normalized = TRUE)
```

### Merge replicates by mean → EP-GFP and NT

```{r counts293_merge}
counts293_merged <- data.frame(
  `EP-GFP` = rowMeans(counts293_norm[, c("EP_1", "EP_2")]),
  NT        = rowMeans(counts293_norm[, c("NT_1", "NT_2")]),
  row.names   = rownames(counts293_norm),
  check.names = FALSE
)
```

### log2(x+1) transform and row-center

```{r counts293_log2center}
# log2(normalized_count + 1) — mirrors BLaER preprocessing
counts293_log2 <- log2(counts293_merged + 1)

# Row-center: subtract each gene's mean across conditions
counts293_row_means <- rowMeans(counts293_log2)
counts293_log2c     <- sweep(counts293_log2, 1, counts293_row_means, FUN = "-")
```

### Annotate with gene symbols

```{r counts293_symbols}
cat("=== 293 symbol annotation ===\n")
counts293_log2centered <- annotate_and_deduplicate(counts293_log2c, numeric_cols = c("EP-GFP", "NT"))

head(counts293_log2centered)
dim(counts293_log2centered)
```

---

## 4. Summary

```{r summary_table}
summary_df <- data.frame(
  Dataset       = c("BLaER", "RCH", "293"),
  Normalization = c(
    "DESeq2 median-of-ratios (pre-computed)",
    "DESeq2 median-of-ratios → log2(x+1)",
    "DESeq2 median-of-ratios → log2(x+1)"
  ),
  Samples = c(
    paste(colnames(blaer_log2centered),     collapse = " | "),
    paste(colnames(rch_log2centered),       collapse = " | "),
    paste(colnames(counts293_log2centered), collapse = " | ")
  ),
  Genes = c(
    nrow(blaer_log2centered),
    nrow(rch_log2centered),
    nrow(counts293_log2centered)
  )
)

knitr::kable(summary_df, caption = "Final log2-centered matrices annotated with gene symbols")
```

---

## 5. Optional: Save outputs

```{r save, eval=FALSE}
write.table(blaer_log2centered,     "blaer_log2centered.txt",     sep = "\t", quote = FALSE)
write.table(rch_log2centered,       "rch_log2centered.txt",       sep = "\t", quote = FALSE)
write.table(counts293_log2centered, "counts293_log2centered.txt", sep = "\t", quote = FALSE)
```

---

## 6. Intersection with TADs

```{r TAD Boundary Gene Analysis, fig.width=14, fig.height=8}
cat("\n=== TAD BOUNDARY GENE ANALYSIS ===\n\n")

# ----------------------------------------------------------------------------
# 1. Load TAD boundary BED files
# ----------------------------------------------------------------------------
tad_dir <- "R:/R_STIK/Hi-C/Xavi_TADs/datasets_comparison"   # same as root.dir in setup

tad_files <- list(
  TP_plus  = list(file = file.path(tad_dir, "boundaries_TP+_all3datasets.bed"),  label = "TP+"),
  TP_minus = list(file = file.path(tad_dir, "boundaries_TP-_all3datasets.bed"),  label = "TP-"),
  shared   = list(file = file.path(tad_dir, "boundaries_shared_all3datasets.bed"), label = "shared")
)

tad_gr_list <- list()

for (tad_name in names(tad_files)) {
  bed <- read.table(
    tad_files[[tad_name]]$file,
    header = FALSE, sep = "\t", stringsAsFactors = FALSE,
    col.names = c("chr", "start", "end", "score", "datasets")
  )
  gr <- makeGRangesFromDataFrame(
    bed %>% mutate(tad_category = tad_files[[tad_name]]$label),
    seqnames.field     = "chr",
    keep.extra.columns = TRUE
  )
  tad_gr_list[[tad_name]] <- gr
  cat(paste0(tad_files[[tad_name]]$label, ": ", length(gr), " boundaries\n"))
}

tad_gr_all <- do.call(c, unname(tad_gr_list))

# ----------------------------------------------------------------------------
# 2. Load gene annotation (gencode v28 BED → gene symbols + coordinates)
# ----------------------------------------------------------------------------
genes_bed <- read.table(
  "R:/R_STIK/xavi_scripts/genes_gencode.v28.bed",
  sep = "\t", header = FALSE,
  col.names = c("chrom", "start", "end", "strand", "gene_name", "ID", "description")
) %>%
  select(chrom, start, end, gene_name) %>%
  mutate(start = as.integer(start), end = as.integer(end))

genes_gr <- makeGRangesFromDataFrame(genes_bed,
                                     seqnames.field     = "chrom",
                                     keep.extra.columns = TRUE)

# ----------------------------------------------------------------------------
# 3. Overlap TAD boundaries → gene symbols
# ----------------------------------------------------------------------------
hits <- findOverlaps(tad_gr_all, genes_gr)

tad_gene_map <- data.frame(
  tad_category = tad_gr_all$tad_category[queryHits(hits)],
  gene_name    = genes_gr$gene_name[subjectHits(hits)]
) %>% distinct()

cat("\nUnique genes per TAD category:\n")
print(table(tad_gene_map$tad_category))
cat("\n")

# ----------------------------------------------------------------------------
# 4. Convert each log2-centered matrix to long format
#    Rownames are HGNC symbols in all three datasets
# ----------------------------------------------------------------------------
to_long <- function(mat, dataset_label) {
  as.data.frame(mat) %>%
    tibble::rownames_to_column("gene_name") %>%
    pivot_longer(
      cols      = -gene_name,
      names_to  = "condition",
      values_to = "log2_centered"
    ) %>%
    mutate(dataset = dataset_label)
}

counts_long_all <- bind_rows(
  to_long(blaer_log2centered,     "BLaER"),
  to_long(rch_log2centered,       "RCH"),
  to_long(counts293_log2centered, "293")
)

# ----------------------------------------------------------------------------
# 5. Join TAD gene map with expression data (across all three datasets)
# ----------------------------------------------------------------------------
tad_category_levels <- c("TP+", "TP-", "shared")

dataset_levels <- c("BLaER", "RCH", "293")

counts_tad <- tad_gene_map %>%
  inner_join(counts_long_all, by = "gene_name") %>%
  mutate(
    tad_category = factor(tad_category, levels = tad_category_levels),
    dataset      = factor(dataset, levels = dataset_levels)
  )

cat("Genes per dataset × TAD category matched in expression matrices:\n")
print(counts_tad %>% distinct(dataset, tad_category, gene_name) %>% dplyr::count(dataset, tad_category))
cat(paste0("\nTotal observations: ", nrow(counts_tad), "\n\n"))

# ----------------------------------------------------------------------------
# 6. Summary statistics
# ----------------------------------------------------------------------------
summary_stats_tad <- counts_tad %>%
  group_by(dataset, tad_category, condition) %>%
  summarise(
    n_genes    = n_distinct(gene_name),
    median_exp = median(log2_centered, na.rm = TRUE),
    mean_exp   = mean(log2_centered,   na.rm = TRUE),
    sd_exp     = sd(log2_centered,     na.rm = TRUE),
    .groups    = "drop"
  )

cat("=== SUMMARY STATISTICS ===\n")
print(summary_stats_tad)
write.table(summary_stats_tad,
            "tad_boundary_counts_summary.tsv",
            sep = "\t", row.names = FALSE, quote = FALSE)

# ----------------------------------------------------------------------------
# Wilcoxon: TP+ vs shared and TP- vs shared (shared as reference)
# Per-gene means per condition per dataset
# ----------------------------------------------------------------------------
gene_means_tad <- counts_tad %>%
  group_by(dataset, tad_category, gene_name, condition) %>%
  summarise(mean_log2 = mean(log2_centered), .groups = "drop")

sig_label <- function(p) {
  dplyr::case_when(
    is.na(p)  ~ "ns",
    p < 0.001 ~ "***",
    p < 0.01  ~ "**",
    p < 0.05  ~ "*",
    TRUE      ~ "ns"
  )
}

wilcox_tad <- gene_means_tad %>%
  group_by(dataset, condition) %>%
  summarise(
    p_TPplus_vs_shared = {
      x <- mean_log2[tad_category == "TP+"]
      y <- mean_log2[tad_category == "shared"]
      if (length(x) >= 3 & length(y) >= 3) wilcox.test(x, y)$p.value else NA_real_
    },
    p_TPminus_vs_shared = {
      x <- mean_log2[tad_category == "TP-"]
      y <- mean_log2[tad_category == "shared"]
      if (length(x) >= 3 & length(y) >= 3) wilcox.test(x, y)$p.value else NA_real_
    },
    .groups = "drop"
  ) %>%
  group_by(dataset) %>%
  mutate(
    padj_TPplus_vs_shared  = p.adjust(p_TPplus_vs_shared,  method = "BH"),
    padj_TPminus_vs_shared = p.adjust(p_TPminus_vs_shared, method = "BH"),
    sig_TPplus             = sig_label(padj_TPplus_vs_shared),
    sig_TPminus            = sig_label(padj_TPminus_vs_shared)
  ) %>%
  ungroup()

cat("\n--- Wilcoxon: TP+ vs shared and TP- vs shared (BH-adjusted within dataset) ---\n")
print(wilcox_tad)
write.table(wilcox_tad,
            "wilcoxon_tad_vs_shared.tsv",
            sep = "\t", row.names = FALSE, quote = FALSE)

# ----------------------------------------------------------------------------
# Significance labels: above TP+ and TP- boxes per facet
# ----------------------------------------------------------------------------
y_positions_tad <- counts_tad %>%
  group_by(dataset, condition, tad_category) %>%
  summarise(y_max = max(log2_centered, na.rm = TRUE), .groups = "drop")

bracket_tad <- wilcox_tad %>%
  pivot_longer(
    cols      = c(sig_TPplus, sig_TPminus),
    names_to  = "which",
    values_to = "sig"
  ) %>%
  filter(sig != "ns") %>%
  mutate(
    tad_category = factor(
      ifelse(which == "sig_TPplus", "TP+", "TP-"),
      levels = tad_category_levels
    )
  ) %>%
  left_join(y_positions_tad, by = c("dataset", "condition", "tad_category")) %>%
  mutate(y_pos = y_max * 1.08)

# ----------------------------------------------------------------------------
# Plot — x = tad_category, facet_grid(dataset ~ condition)
# ----------------------------------------------------------------------------
p_tad <- ggplot(
  counts_tad,
  aes(x = tad_category, y = log2_centered, fill = tad_category)
) +
  geom_boxplot(
    outlier.size  = 0.4,
    outlier.alpha = 0.3,
    alpha         = 0.7,
    linewidth     = 0.4
  ) +
  geom_hline(
    yintercept = 0,
    linetype   = "dashed",
    color      = "gray30",
    linewidth  = 0.4
  ) +
  {if (nrow(bracket_tad) > 0)
    geom_text(
      data        = bracket_tad,
      aes(x       = tad_category, y = y_pos, label = sig),
      size        = 5,
      fontface    = "bold",
      color       = "gray20",
      inherit.aes = FALSE
    )
  } +
  scale_fill_manual(values = c("TP+" = "#d6604d", "TP-" = "#4393c3", "shared" = "#878787")) +
  facet_grid(dataset ~ condition, scales = "free_y") +
  theme_bw() +
  theme(
    axis.text.x      = element_text(angle = 45, hjust = 1, size = 10),
    axis.title       = element_text(size = 11),
    plot.title       = element_text(hjust = 0.5, face = "bold", size = 13),
    strip.text       = element_text(face = "bold", size = 11),
    strip.background = element_rect(fill = "grey95", color = "black"),
    legend.position  = "none",
    panel.grid.minor = element_blank()
  ) +
  xlab("TAD boundary category") +
  ylab("Log2-centered expression") +
  labs(
    title    = "Gene expression at TAD boundaries across datasets",
    subtitle = "Significance labels: TP+ or TP− vs shared (Wilcoxon, BH-adjusted)"
  )

print(p_tad)
ggsave("tad_boundary_counts_vs_shared_boxplot.pdf", p_tad, width = 12, height = 10)
```

```{r TAD Boundary TP Gene Lists}
cat("\n=== GENE LISTS AT TP+ AND TP- BOUNDARIES ===\n\n")

# ----------------------------------------------------------------------------
# For each dataset, retrieve all genes overlapping TP+ and TP- boundaries
# with their expression values in all conditions
# ----------------------------------------------------------------------------

tp_gene_lists <- list()

for (ds in dataset_levels) {
  for (cat in c("TP+", "TP-")) {
    
    label <- paste0(ds, "_", cat)
    
    # Wide format: one row per gene, one column per condition
    genes_tp <- gene_means_tad %>%
      filter(dataset == ds, tad_category == cat) %>%
      pivot_wider(
        id_cols     = c(dataset, tad_category, gene_name),
        names_from  = condition,
        values_from = mean_log2
      ) %>%
      arrange(gene_name)
    
    cat(paste0(label, ": ", nrow(genes_tp), " genes\n"))
    tp_gene_lists[[label]] <- genes_tp
    
    write.table(
      genes_tp,
      paste0("genes_", label, ".tsv"),
      sep = "\t", row.names = FALSE, quote = FALSE
    )
  }
}

# ----------------------------------------------------------------------------
# Combined file: all TP+ and TP- genes across all datasets (long format)
# ----------------------------------------------------------------------------
tp_genes_combined <- gene_means_tad %>%
  filter(tad_category %in% c("TP+", "TP-")) %>%
  arrange(dataset, tad_category, gene_name)

write.table(
  tp_genes_combined,
  "genes_TP_boundaries_all_datasets.tsv",
  sep = "\t", row.names = FALSE, quote = FALSE
)

cat(paste0("\nTotal unique genes across TP+ and TP- (any dataset):\n"))
tp_genes_combined %>%
  distinct(dataset, tad_category, gene_name) %>%
  dplyr::count(dataset, tad_category) %>%
  print()

cat("\nUnique gene symbols overlapping TP+ or TP- in at least one dataset:\n")
tp_genes_combined %>%
  distinct(gene_name) %>%
  nrow() %>%
  cat("\n")
```